---
title: "Loci sample size limits for snaq"
author: 
    - Gustavo A. Ballen^[Museu de Zoologia da Universidade de São Paulo]
output: 
    pdf_document:
        fig_caption: yes
        number_sections: true
        citation_package: natbib
        latex_engine: xelatex
urlcolor: blue
toc: true
biblio-style: apalike
bibliography: /home/balleng/Dropbox/Gustavo/Bibtex/PhyloNetworks.bib
header-includes: 
- \usepackage{lscape}
- \usepackage{longtable}
---

```{r setup, echo = FALSE}
library(knitr)
knit_hooks$set(small.mar = function(before, options, envir) {
    if (before)    par(mar=c(5,4,1,1)+.1)  # smaller margin on top and right
})
knitr::opts_chunk$set(tidy = TRUE, small.mar = TRUE, comment = "")
```

```{r}
# load R packages
library(ape)
```

# Goal

We want to understand the limits of snaq with respect to sample size in terms of loci. \citet{Solis-Lemus2017} used simulations for exploring how data-hungry is SNaQ and found that for moderate-sized networks (say 15 leaves) there were necessary hundreds (and maybe thousands) of loci from phylogenomic datasets. However, these sample size limits are not well understood and probably will depend on factors such as hybridization morphology (e.g., how many nodes are involved in the cycle, the more, the easier to detect), the nature of the inheritance probabilities (e.g., whether lambda on each hybrid vertex is equal 0.5, 0.5, or heterogeneous, and whether strongly heterogeneous settings affect sample size needed for detection), and network size as measured in terms of number of leaves.

We are going to consider here level-1 networks with three levels of cycle complexity: 3, 4, and 6 nodes. Finally, we will consider three settings of lambda values: homogeneous hybridization (0.5, 0.5), moderately heterogeneous (0.75, 0.25), and drastically heterogeneous (0.05, 0.95).

# Simulating loci on a network

First create output directories for each of the analyses

```{bash, eval = FALSE}
mkdir net_h1_n6_ho
mkdir hybridLambda_output
mkdir bucky_output
mkdir snaq_output
```

Use this arbitrary tree:

```
((((A,B),(C,D)),(E,(F,G))),H);
```

Construct a hybrid edge of six nodes. Pick two edges that will form a cycle with six nodes, and manually input the hybryd nodes. That generates the network below:

```
(((((A,B))#H,(C,D)),(E,(F,(G,#H)))),H);
```

or in hybridLambda format (including a $\gamma = 0.5$):

```
(((((A,B))H#0.5,(C,D)),(E,(F,(G,H#0.5)))),H);
```

Now use hybridLambda for simulating 3000 loci on a given tree, this time with one hybridization, six nodes in the cycle, and homogeneous gamma:

```{bash}
hybrid-Lambda -spcu '(((((A,B))H#0.5,(C,D)),(E,(F,(G,H#0.5)))),H);' -num 3000 -seed 1985 -o 'hybridLambda_output/net_h1_n6_ho'
```

Check results:

```{bash}
tree .
```

Apparently the function `readMultiTopology` from PhyloNetworks do not like the gene tree format from hybridLambda. The reason is that hybridLambda is converting a given hybrid edge into a tip called e.g., H#0.5_1. The message `Expected name after # but received 0 in left parenthesis 4...` suggests that `#` is being interpreted as a special character for defining a hybrid node in extnewick. Thus, `sed`ing it into a string (say `hyb`) for reference should suffice:

```{bash}
# backup the original results
cp hybridLambda_output/net_h1_n6_ho_coal_unit hybridLambda_output/net_h1_n6_ho_coal_unit.bak

# replace the numeral with the string 'hyb'
sed -i 's/#/hyb/g' hybridLambda_output/net_h1_n6_ho_coal_unit
```

Now use the 3000 gene trees as input for SNaQ and use the theoretical tree as starting tree:

```{julia, eval = FALSE}
### h1_n6_ho_net0.jl
using Distributed
addprocs(4)
@everywhere using PhyloNetworks

using PhyloPlots
using CSV

# load input trees
trees = readMultiTopology("hybridLambda_output/net_h1_n6_ho_coal_unit")
# terminals must be renamed in order to match gene-tree names from hybridLambda
# also, an arbitrary position for the hybrid-induced terminal was chosen to be sister to G
start_tree = readTopology("((((A_1,B_1),(C_1,D_1)),(E_1,(F_1,(G_1,Hhyb0.5_1)))),H_1);")

# calculate CFs
q,t = countquartetsintrees(trees)
df = writeTableCF(q,t)
cfs = readTableCF(df)

# calculate a h=0 network
net0 = snaq!(start_tree, cfs, hmax=0, filename="h1_n6_ho_net0", seed=1234)
```

A transcript of the output is:

```{bash, eval = FALSE}
julia h1_n6_ho_net0.jl 2>&1 h1_n6_h0_net0.outerr 
Reading in trees, looking at 126 quartets in each...
0+--------------------------------------------------+100%
  **************************************************
3000.0 gene trees per 4-taxon set
optimization of topology, BL and inheritance probabilities using:
 hmax = 0,
 tolerance parameters: ftolRel=1.0e-6, ftolAbs=1.0e-6,
                       xtolAbs=0.001, xtolRel=0.01.
 max number of failed proposals = 75, liktolAbs = 1.0e-6.
rootname for files: h1_n6_ho_net0
BEGIN: 10 runs on starting tree (A_1,B_1,((C_1,D_1):0.0,((E_1,(F_1,(G_1,Hhyb0.5_1):0.002503130218118477):0.0):0.008578355774871729,H_1):0.0):0.011313759900273404);
       using 5 processors
2021-06-11 0:14:21.516
      From worker 3:	seed: 1234 for run 1, 2021-06-11 0:14:37.401
      From worker 5:	seed: 76679 for run 3, 2021-06-11 0:14:37.455
      From worker 4:	seed: 59084 for run 2, 2021-06-11 0:14:37.532
      From worker 2:	seed: 56623 for run 4, 2021-06-11 0:14:37.627
      From worker 3:	seed: 46008 for run 5, 2021-06-11 0:17:25.128
      From worker 5:	seed: 79402 for run 6, 2021-06-11 0:17:43.451
      From worker 2:	seed: 85414 for run 7, 2021-06-11 0:17:49.578
      From worker 4:	seed: 20058 for run 8, 2021-06-11 0:18:41.878
      From worker 2:	seed: 29861 for run 9, 2021-06-11 0:19:51.07
      From worker 3:	seed: 24683 for run 10, 2021-06-11 0:20:3.762

MaxNet is (((((B_1,E_1):0.0028841945508996914,Hhyb0.5_1):0.0035572252369179057,F_1):0.005254479335434281,(C_1,G_1):0.007742950671429925):0.0011738626208142714,(H_1,A_1):0.014900497457803537,D_1); 
with -loglik 2.3670289296421987
```
