---
title: "Networks characterization from Birth Dead method with NetSim"
author: 
    - Carlos Acosta
output: 
  bookdown::pdf_document2:
  fig_caption: yes
  number_sections: true
urlcolor: blue
toc: true
fig_caption: yes

header-includes:
- \usepackage{float}
---

```{r setup, echo = FALSE}
#library(kableExtra)
#library("bookdown")
library(knitr)
knit_hooks$set(small.mar = function(before, options, envir) {
    if (before)    par(mar=c(5,4,1,1)+.1)  # smaller margin on top and right
})
knitr::opts_chunk$set(tidy = TRUE, small.mar = TRUE, comment = "")
```

# Packages

```{r,echo=TRUE}
library(NetSim) # simulate bd time-tree networks: https://github.com/jjustison/NetSim
library(TreeSim) # simulate bd time-trees
library(ape) # tools for handling phylo objects
library(diversitree) # tools for estimating lambda and mu
library(phytools)
library("dplyr")    # Data manipulation
library(ggplot2)
library(corrplot)
library(cluster)
```

# Using NetSim:

Simulation of 300 Trees with the following parameters using NetSim:

* lambda = 0.9
* mu = 0.5
* nu = 0.03
* hybprops = c(0.5, 0.5, 0.5)

```{r chunk, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10)}
set.seed(500)
numbsim=300
bdNS <- NetSim::sim.bdh.taxa.ssa(n=10, 
                                 numbsim = numbsim, 
                                 lambda = 0.9, 
                                 mu = 0.5, 
                                 nu = 0.03,
                                 hybprops = c(0.5, 0.5, 0.5), 
                                 hyb.inher.fxn = NetSim::make.beta.draw(1, 1), 
                                 frac = 1,
                                 mrca = FALSE, 
                                 complete = TRUE, 
                                 stochsampling = FALSE, 
                                 hyb.rate.fxn = NULL, 
                                 trait.model = NULL)
```

As we can see in  figure \@ref(fig:chunk1), 59% of the trees are extinct, 19.67% are not reticulation and only 21.33% are reticulated tree.

```{r chunk1, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10),fig.cap = "\\label{fig:chunk1} Percentage of tree types using NetSim with parameters lambda = 0.9, mu = 0.5 and nu = 0.03,"}
#Code to plot 
Summ<-summary(bdNS)[,1]
datSum<-data.frame(Var=Summ,Freq=1)
datSum2<-aggregate(Freq~Var,FUN="sum",data=datSum)
datSum2[,1]<-as.character(datSum2[,1])
datSum2[datSum2[,1]==1,1]<-"Extint Tree"
datSum2[datSum2[,1]==6,1]<-"Not Reticulation Tree"
datSum2[datSum2[,1]==7,1]<-"Reticulation Tree"
datSum2$Prop<-datSum2[,2]/sum(datSum2[,2])

#codes to make a nice pie with ggplot2
colnames(datSum2)<-c("class","n","prop")
count.data<-datSum2

# Add label position
count.data <- count.data %>%
    arrange(desc(class)) %>%
    mutate(lab.ypos = cumsum(prop) - 0.5*prop)
count.data$Por<-round(count.data$prop*100,2)

ggplot(data = count.data, 
       aes(x = 2, y = prop, fill = class))+
  geom_bar(stat = "identity")+
  coord_polar("y", start = 200) +
  geom_text(aes(y = lab.ypos, label = paste(Por,"%", sep = "")), col = "white") +
  theme_void() +
  scale_fill_brewer(palette = "Dark2")+
  xlim(.2,2.5)

```

## Extraction of information from tree
We use the tree 35 (see fig. \@ref(fig:chunk4)) with three reticulations to extract the tree information, and then extend of all the Trees. We extract and analice the following information:

* Mean of edge length (meEdgeL)
* cv of edge length (cvEdgeL)
* Number of tips (TipNum)
* Proportions of extintct taxa (PorExtin)
* Number of internal nodes (NodeNum)
* Time to reach 10 taxa (TimeTree)
* Reticulation Number (RetNum)
* Average time of birth of hybrids (TSHmean)
* Minimum time of birth of hybrids (TSHmin)
* Maximun time of birth of hybrids (TSHmax)

If reticulation number is one, TSHmean, TSHmin and TSHmax are the same, if there are not reticulation the value is cero.

```{r chunk3, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10),fig.cap = "\\label{fig:chunk4} Tree 35 which have three reticulations"}
## remove any trees with no taxa
bdNS <- bdNS[!sapply(X = bdNS, FUN = is.null)]

#Information about tree goes extinct=0 and no extinct tips are sampled=1
bdNS <- bdNS[sapply(X = bdNS, FUN = is.phylo)]#birth dead tree with NetSim with value 0 or 1
Tree2<-bdNS[[35]]#15#10#96
plot(Tree2)
#Tree2
```

```{r chunk4.1, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10)}
###Extracting general Information
meEdgeL<-mean(Tree2$"edge.length")#mean of edge length
sdEdgeL<-var(Tree2$"edge.length")
cvEdgeL<-sdEdgeL/meEdgeL#cv of edge length
TipNum<-length(Tree2$"tip.label")# Number of tips
PorExtin<-(TipNum-10)/TipNum#Proportions of extintct taxa 
NodeNum<-Tree2$Nnode#Number of internal Nodes# Maybe we can also divide it for TipNum

LT2<-ltt(Tree2,plot=FALSE)
Tim<-LT2$times
TimeTree<-max(Tim)#Time to reach 10 taxa

#extracting information about the Reticulations

Ret<-Tree2$"reticulation"
Ret1<-Ret
RetNum<-nrow(Ret)# Reticulation Number

if(RetNum>=1){
for(i in 1:nrow(Ret1)){
for(j in 1:ncol(Ret1)){
Ret1[i,j]<-Tim[names(Tim)%in%Ret1[i,j]]
}
}

TimeStartHibr<-apply(Ret1,1,max)/TimeTree#Time when birth the hibirds, we can divide it for the time to reach 10 taxa (TimeTree)
TSHmean<-mean(TimeStartHibr)# Average time of birth of hybrids 
TSHmin<-min(TimeStartHibr)# minimum time of birth of hybrids 
TSHmax<-max(TimeStartHibr)# maximun time of birth of hybrids 
}else{
TSHmean<-0
TSHmin<-0
TSHmax<-0
}

#If reticulation number is one, TSHmean, TSHmin and TSHmax are the same, if there are not reticulation the value is cero.
dat<-data.frame(TipNum,NodeNum,PorExtin,meEdgeL,cvEdgeL,TimeTree,RetNum,TSHmean,TSHmin,TSHmax)

dat
```


## Extention of the analisys in all the trees
We extended the analisys performed in section 2.1 on all trees

```{r chunk4, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10)}
out1<-c()
for(k in 1:length(bdNS)){
Tree2<-bdNS[[k]]
###Extracting general Information
meEdgeL<-mean(Tree2$"edge.length")#mean of edge length
sdEdgeL<-var(Tree2$"edge.length")
cvEdgeL<-sdEdgeL/meEdgeL#cv of edge length
TipNum<-length(Tree2$"tip.label")# Number of tips
PorExtin<-(TipNum-10)/TipNum#Proportions of extintct taxa 
NodeNum<-Tree2$Nnode#Number of internal Nodes# Maybe we can also divide it for TipNum

LT2<-ltt(Tree2,plot=FALSE)
Tim<-LT2$times
TimeTree<-max(Tim)#Time to reach 10 taxa

#extracting information about the Reticulations

Ret<-Tree2$"reticulation"
Ret1<-Ret
RetNum<-nrow(Ret)# Reticulation Number

if(RetNum>=1){
for(i in 1:nrow(Ret1)){
for(j in 1:ncol(Ret1)){
Ret1[i,j]<-Tim[names(Tim)%in%Ret1[i,j]]
}
}

TimeStartHibr<-apply(Ret1,1,max)/TimeTree#Time when birth the hibirds, we can divide it for the time to reach 10 taxa (TimeTree)
TSHmean<-mean(TimeStartHibr)# Average time of birth of hybrids 
TSHmin<-min(TimeStartHibr)# minimum time of birth of hybrids 
TSHmax<-max(TimeStartHibr)# maximun time of birth of hybrids 
}else{
TSHmean<-0
TSHmin<-0
TSHmax<-0
}

#If reticulation number is one, TSHmean, TSHmin and TSHmax are the same, if there are not reticulation the value is cero.
dat<-data.frame(TipNum,NodeNum,PorExtin,meEdgeL,cvEdgeL,TimeTree,RetNum,TSHmean,TSHmin,TSHmax)

out1<-rbind(out1,dat)
}

head(out1)
```

## Analisys per variable
In this section we analice the Trees varaible per variable (see fig \@ref(fig:chunk5)), as you can see, only PorExtin, meEdgeL, cvEdgeL, and TimeTree have a distribution similar to the normal; NodeNum has aHalf-normal distribution.

```{r chunk63, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10),fig.cap = "\\label{fig:chunk5} Univariate plot for all the information extracted"}
par(mfrow=c(3,3))
nam<-colnames(out1)
for(i in c(2:6,8:10)){
  hist(out1[,i],main=nam[i])
}
barplot(table(sort(out1$RetNum)),main="Reticulation Number")
```

## Multivariate Analisys
In this part we perform a cor plot (see fig\@ref(fig:chunk64)), in which we can see that Reticulation number is positively correlated with NodeNum. Besides we make a cluster analisys whit euclidean distance and UPGMA grouped method(see fig \@ref(fig:chunk7)), which sugest two cluster numbers, so we can choose one to analyze with shaq. 

```{r chunk73, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10),fig.cap = "\\label{fig:chunk64} Corelation plot",echo=TRUE}

cor.mtest <- function(mat, ...) {
mat <- as.matrix(mat)
n <- ncol(mat)
p.mat<- matrix(NA, n, n)
diag(p.mat) <- 0
for (i in 1:(n - 1)) {
for (j in (i + 1):n) {
tmp <- cor.test(mat[, i], mat[, j], ...)
p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
}
}
colnames(p.mat) <- rownames(p.mat) <- colnames(mat)
p.mat
}


M<-cor(out1)
p.mat <- cor.mtest(out1)
col <- colorRampPalette(c("#BB4444", "#EE9988", "#FFFFFF", "#77AADD", "#4477AA"))
corrplot(M, method="color", col=col(200),
type="upper", order="hclust",
addCoef.col = "black", # Add coefficient of correlation
tl.col="black", tl.srt=45, #Text label color and rotation
# Combine with significance
p.mat = p.mat, sig.level = 0.01, insig = "blank",
# hide correlation coefficient on the principal diagonal
diag=FALSE
)

```




```{r chunktr73, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10),fig.cap = "\\label{fig:chunk7} Cluster Analisys"}
clust<-agnes(out1, metric = "euclidean",stand = TRUE,method="average")

par(mfrow=c(1,2))
plot(clust)
```
