---
title: "Networks characterization from Birth Dead method with NetSim"
author: 
    - Carlos Acosta
output: 
  bookdown::pdf_document2:
  fig_caption: yes
  number_sections: true
urlcolor: blue
toc: true
fig_caption: yes

header-includes:
- \usepackage{float}
---

```{r setup, echo = FALSE}
#library(kableExtra)
#library("bookdown")
library(knitr)
knit_hooks$set(small.mar = function(before, options, envir) {
    if (before)    par(mar=c(5,4,1,1)+.1)  # smaller margin on top and right
})
knitr::opts_chunk$set(tidy = TRUE, small.mar = TRUE, comment = "")
```

# Packages

```{r,echo=TRUE}
library(NetSim) # simulate bd time-tree networks: https://github.com/jjustison/NetSim
library(TreeSim) # simulate bd time-trees
library(ape) # tools for handling phylo objects
library(diversitree) # tools for estimating lambda and mu
library(phytools)
library("dplyr")    # Data manipulation
library(ggplot2)
library(corrplot)
library(cluster)
library(vioplot)
library(geiger)
```

# Functions
## Functions with reticulations not and recornozible by snaq and not level-1 reticulation
```{r chunk1, cache = TRUE, echo=TRUE}

#We want to know if the reticulation node produce a Tip or Node
##This function goes in the direction of the tips omiting reticulations nodes until it finds one no reticulation node
DesJupRet<-function(ED,reff,father,allret){
ti12<-reff
Log=TRUE
while(Log){
ti12<-ED[ED[,1]==ti12&ED[,2]!=father,2][1]
#we use father because when we want to evaluate RootRet we want to know if [Tip or Node 3] is a node or tip, no go for the reticulation way.
Log<-ti12%in%allret
}
return(ti12)
}


#This function goes in the direction of the tips jumping node by node and determinate if the refference (reff) produce a Tip or not
Tip<-function(ED,reff,father,allret){
Log=TRUE
i=0
while(Log){
reff<-DesJupRet(ED=ED,reff=reff,father=father,allret=allret)
#
Log<-!(i==2|is.na(reff))
i=i+1
}
if(i==2){res<-"Tip"}else if(i==3){res<-"NoTip"}
return(res)
}


##This function goes in the direction of the root omiting reticulations nodes until it finds one no reticulation node
AssJupRet<-function(ED,reff,allret){
ti12<-reff
Log=TRUE
while(Log){
ti12<-ED[ED[,2]==ti12,1][1]###Loop
Log<-ti12%in%allret
}
return(ti12)
}

#This function goes in the direction of the root jumping node by node and determinate if the refference (reff) produce a root or node
AssRoot<-function(ED,reff,allret){
Log=TRUE
i=0
while(Log){
reff<-AssJupRet(ED=ED,reff=reff,allret=allret)
Log<-!(i==2|is.na(reff))
i=i+1
}
if(i==2){res<-"Root"}else if(i==3){res<-"NoRoot"}else if (i<=1){res<-"No-Identificable"}
return(res)
}

#This function use the functions described before and determinate if the sibling reticulation can be Recognozible by snaq
Identiff<-function(Tree2,Hib1,Hib2,ED,father,RootTest,allret){
Tiprev<-Tip(ED,reff=RootTest,father=father,allret=allret)
Roo<-AssRoot(ED,reff=RootTest,allret=allret)
if(any(Roo%in%c("No-Identificable",NA,NULL))){resAll<-"No-Identificable"}else{

NodTip1<-Tiprev=="NoTip"|Roo=="NoRoot"
NodTip2<-Tip(ED,reff=Hib1,father=father,allret=allret)=="NoTip"
NodTip3<-Tip(ED,reff=Hib2,father=father,allret=allret)=="NoTip"
res1<-sum(c(NodTip1,NodTip2,NodTip3),na.rm = TRUE)#using (+) whit NA or NULL error will be apear, but if we use sum(,na.rm = TRUE), we omith that.7
#res1<-NodTip1+NodTip2+NodTip3
if(res1>=2){resAll<-"Identificable"}else {resAll<-"No-Identificable"} }
return(resAll)
}


# This function analize the tree, determinate if there are reticulations and in the case that it exists evalaute if the reticulation is among siblings, and in the case that it exists evaluate if it is or not recognozible by snaq.

SibCross<-function(Tree2){
RetEv<-Tree2$"reticulation"
allret<-unique(as.numeric(RetEv))
ED<-Tree2$edge

if(nrow(RetEv)==0){SiblingCross<-"NotRet"
res<-list(infor=SiblingCross,ret=NA)}else{
outRet<-c()

for(h in 1:nrow(RetEv)){
RetEv1<-RetEv[h,]
Hib1<-RetEv1[2]#Hibrid
Hib2<-RetEv1[1]#Father
ED<-Tree2$edge

Or1whitoutRet<-AssJupRet(ED,Hib1,allret)#Find the father omiting the reticulations
Or2whitoutRet<-AssJupRet(ED,Hib2,allret)#Find the father omiting the reticulations

Orig1<-ED[ED[,2]==Hib1,1]#Hibrid
Orig2<-ED[ED[,2]==Hib2,1]#Father
 
res<-Orig1==Orig2

if(any(c(Hib1,Hib2)%in%c(Orig1,Orig2))){SiblingCross<-"RetCiclicI"}else if(Or1whitoutRet==Or2whitoutRet&Orig1==Orig2){
father<-Orig1
RootTest<-AssJupRet(ED,Orig1,allret)#Go direction to the root avoiding the reticulation nodes.

Ident<-Identiff(Tree2,Hib1=Hib1,Hib2=Hib2,ED,father,RootTest,allret=allret)
SiblingCross<-Ident
}else if(Or1whitoutRet==Or2whitoutRet&Orig1!=Orig2){SiblingCross<-"not level-1"}else if(Or1whitoutRet!=Or2whitoutRet&Orig1!=Orig2){SiblingCross<-"Identificable"}#"NoSibCross"

outRet<-c(outRet,SiblingCross)
}
res<-list(infor=outRet,ret=RetEv)
}

return(res)
}

```

## Function to analize Network

```{r chunk2, cache = TRUE, echo=TRUE}
RetAnalysis<-function(TreeAnalize1){
mat<-matrix(0,ncol=3,nrow=1)
colnames(mat)<-c("NotIdent", "Ident", "NotLevel-1","RetCiclicI")
colnam<-c("No-Identificable","Identificable","not level-1","RetCiclicI")

#plot(TreeAnalize1)
sibcro1<-SibCross(TreeAnalize1)
tab1<-table(sibcro1$infor)
namtab<-names(tab1)
for(j in 1:length(namtab)){
mat[,colnam==namtab[j]]<-tab1[j]
}
return(mat)
}






NetDescription<-function(bdNS){
out1<-c()
for(k in 1:length(bdNS)){
Tree2<-bdNS[[k]]
###Extracting general Information
meEdgeL<-mean(Tree2$"edge.length")#mean of edge length
sdEdgeL<-var(Tree2$"edge.length")
cvEdgeL<-sdEdgeL/meEdgeL#cv of edge length
#TipNum<-length(Tree2$"tip.label")# Number of tips
#PorExtin<-(TipNum-10)/TipNum#Proportions of extintct taxa 
NodeNum<-Tree2$Nnode#Number of internal Nodes# Maybe we can also divide it for TipNum
LT2<-ltt(Tree2,plot=FALSE)
Tim<-LT2$times
TimeTree<-max(Tim)#Time to reach 10 taxa
#extracting information about the Reticulations
Ret<-Tree2$"reticulation"
Ret1<-Ret
RetNum<-nrow(Ret)# Reticulation Number
if(RetNum>=1){
for(i in 1:nrow(Ret1)){
for(j in 1:ncol(Ret1)){
Ret1[i,j]<-Tim[names(Tim)%in%Ret1[i,j]]
}
}
TimeStartHibr<-apply(Ret1,1,max)/TimeTree#Time when birth the hibirds, we can divide it for the time to reach 10 taxa (TimeTree)
TSHmean<-mean(TimeStartHibr)# Average time of birth of hybrids 
TSHmin<-min(TimeStartHibr)# minimum time of birth of hybrids 
TSHmax<-max(TimeStartHibr)# maximun time of birth of hybrids 
}else{
TSHmean<-0
TSHmin<-0
TSHmax<-0
}

SiblingCross<-RetAnalysis(TreeAnalize1=Tree2)


#If reticulation number is one, TSHmean, TSHmin and TSHmax are the same, if there are not reticulation the value is cero.
#dat<-data.frame(TipNum,NodeNum,PorExtin,meEdgeL,cvEdgeL,TimeTree,RetNum,TSHmean,TSHmin,TSHmax)
dat<-data.frame(NodeNum,meEdgeL,cvEdgeL,TimeTree,RetNum,TSHmean,TSHmin,TSHmax,SiblingCross)
out1<-rbind(out1,dat)
}
return(out1)
}


```


# Parameters

Simulation of 300 Trees with the following parameters using NetSim:

* lambda = 0.9
* mu = 0
* nu = 0.01 and 0.03
* hybprops = See table

```{r chunk3, cache = TRUE,echo=TRUE}

#hybprops Combinations
Upp1<-1
Low1<-0
Gen<-c(rep(Upp1,4),rep(Low1,4))
Deg<-rep(c(Upp1,Upp1,Low1,Low1),2)
Neu<-rep(c(Low1,Upp1),4)


HybTab1<-data.frame(Gen,Deg,Neu)
HybTab<-HybTab1[-7,]#No acepta la combinacion c(0,0,0)
HybTab

```

# simulation data

```{r chunk4, cache = TRUE,echo=TRUE}
#################################
set.seed(501)

outputNu0.01<-c()
for(i in 1:nrow(HybTab)){
numbsim=300
n1<-15
bdNS <- NetSim::sim.bdh.taxa.ssa(n=n1, 
                                 numbsim = numbsim, 
                                 lambda = 0.9, 
                                 mu = 0, 
                                 nu = 0.01,
                                 hybprops = HybTab[i,], 
                                 hyb.inher.fxn = NetSim::make.beta.draw(1, 1), 
                                 frac = 1,
                                 mrca = FALSE, 
                                 complete = TRUE, 
                                 stochsampling = FALSE, 
                                 hyb.rate.fxn = NULL, 
                                 trait.model = NULL)
#
bdNS <- bdNS[!sapply(X = bdNS, FUN = is.null)]
#Information about tree goes extinct=0 and no extinct tips are sampled=1
bdNS <- bdNS[sapply(X = bdNS, FUN = is.phylo)]#birth dead tree with NetSim with value 0 or 1

res<-NetDescription(bdNS)
res2<-data.frame(HybTab[i,],res,row.names = NULL)
outputNu0.01<-rbind(outputNu0.01,res2)
}




outputNu0.02<-c()
for(i in 1:nrow(HybTab)){
numbsim=300
n1<-15
bdNS <- NetSim::sim.bdh.taxa.ssa(n=n1, 
                                 numbsim = numbsim, 
                                 lambda = 0.9, 
                                 mu = 0, 
                                 nu = 0.02,
                                 hybprops = HybTab[i,], 
                                 hyb.inher.fxn = NetSim::make.beta.draw(1, 1), 
                                 frac = 1,
                                 mrca = FALSE, 
                                 complete = TRUE, 
                                 stochsampling = FALSE, 
                                 hyb.rate.fxn = NULL, 
                                 trait.model = NULL)
#
bdNS <- bdNS[!sapply(X = bdNS, FUN = is.null)]
#Information about tree goes extinct=0 and no extinct tips are sampled=1
bdNS <- bdNS[sapply(X = bdNS, FUN = is.phylo)]#birth dead tree with NetSim with value 0 or 1

res<-NetDescription(bdNS)
res2<-data.frame(HybTab[i,],res,row.names = NULL)
outputNu0.02<-rbind(outputNu0.02,res2)
}






## reorganize the data
outputNu0.01$CombHybP<-apply(outputNu0.01[,c("Gen", "Deg", "Neu")],1,paste0,collapse="-")
outputNu0.02$CombHybP<-apply(outputNu0.02[,c("Gen", "Deg", "Neu")],1,paste0,collapse="-")


agNu0.02<-aggregate(RetNum~CombHybP,FUN=mean,data=outputNu0.02)
agNu0.02Ord<-agNu0.02[order(agNu0.02$RetNum),]

LevChar<-as.character(agNu0.02Ord$CombHybP)

outputNu0.02[,"CombHybP"]<-factor(outputNu0.02[,"CombHybP"],levels=LevChar)
outputNu0.01[,"CombHybP"]<-factor(outputNu0.01[,"CombHybP"],levels=LevChar)
#########################################
head(outputNu0.02)
head(outputNu0.01)
```


# Reticulation functions
```{r chunk55.1, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10)}
numbsim=500
n1<-15
bdNS <- NetSim::sim.bdh.taxa.ssa(n=n1, 
                                 numbsim = numbsim, 
                                 lambda = 0.9, 
                                 mu = 0, 
                                 nu = 0.03,
                                 hybprops = c(0.5,0.25,0.25), 
                                 hyb.inher.fxn = NetSim::make.beta.draw(1, 1), 
                                 frac = 1,
                                 mrca = FALSE, 
                                 complete = TRUE, 
                                 stochsampling = FALSE, 
                                 hyb.rate.fxn = NULL, 
                                 trait.model = NULL)
#
bdNS <- bdNS[!sapply(X = bdNS, FUN = is.null)]
#Information about tree goes extinct=0 and no extinct tips are sampled=1
bdNS <- bdNS[sapply(X = bdNS, FUN = is.phylo)]

```

## No Recognozible by snaq
```{r chunk55.2, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10)}
#No Identificable for snaq
out2<-c()
for(i in 1:length(bdNS)){
res<-SibCross(bdNS[[i]])$infor
res2<-any(res=="No-Identificable")
out2<-c(out2,res2)
}
posNotIdent<-which(out2);posNotIdent

TreeAnalize<-bdNS[[posNotIdent[1]]]#4
sibcro<-SibCross(TreeAnalize)
nod<-sibcro$ret[sibcro$infor=="No-Identificable",]
plot(TreeAnalize)
nodelabels(node=nod)

TreeAnalize<-bdNS[[posNotIdent[3]]]#4
sibcro<-SibCross(TreeAnalize)
nod<-sibcro$ret[sibcro$infor=="No-Identificable",]
plot(TreeAnalize)
nodelabels(node=nod)

```


## No level One
```{r chunk55.3, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10)}
#No Identificable for snaq
out2<-c()
for(i in 1:length(bdNS)){
res<-SibCross(bdNS[[i]])$infor
res2<-any(res=="not level-1")
out2<-c(out2,res2)
}
posNotIdent<-which(out2);posNotIdent

TreeAnalize<-bdNS[[posNotIdent[2]]]#4
sibcro<-SibCross(TreeAnalize)
nod<-sibcro$ret[sibcro$infor=="not level-1",]
plot(TreeAnalize)
nodelabels(node=nod)

TreeAnalize<-bdNS[[posNotIdent[3]]]#4
sibcro<-SibCross(TreeAnalize)
nod<-sibcro$ret[sibcro$infor=="not level-1",]
plot(TreeAnalize)
nodelabels(node=nod)
nodelabels()
```

## Ciclic reticulations
```{r chunk55.4, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10)}
#No Identificable for snaq
out2<-c()
for(i in 1:length(bdNS)){
res<-SibCross(bdNS[[i]])$infor
res2<-any(res=="RetCiclicI")
out2<-c(out2,res2)
}
posNotIdent<-which(out2);posNotIdent

TreeAnalize<-bdNS[[posNotIdent[2]]]#4
sibcro<-SibCross(TreeAnalize)
nod<-sibcro$ret[sibcro$infor=="RetCiclicI",]
plot(TreeAnalize)
nodelabels(node=nod)

TreeAnalize<-bdNS[[posNotIdent[3]]]#4
sibcro<-SibCross(TreeAnalize)
nod<-sibcro$ret[sibcro$infor=="RetCiclicI",]
plot(TreeAnalize)
nodelabels(node=nod)

```


# Results 

## Number of reticulation
### Barplot
```{r chunk5, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10)}
colors1<-c("#daf8e3",
"#97ebdb",
"#00c2c7",
"#0086ad",
"#005582",

"#ffbaba",
"#ff7b7b",
"#ff5252",
"#ff0000",
"#a70000")



## Number of reticulation

tabbNu0.01<-table(outputNu0.01[,c("RetNum","CombHybP")])
tabbNu0.02<-table(outputNu0.02[,c("RetNum","CombHybP")])

xxlab1<-"hybprops combination                              "
pop<-par(mfrow=c(1,2),oma = c(0, 0, 2, 0))
barplot(tabbNu0.02,legend.text=rownames(tabbNu0.02),col=colors1,xlim=c(0,11),main="nu=0.02",xlab=xxlab1,las=2)
barplot(tabbNu0.01,legend.text=rownames(tabbNu0.01),col=colors1,xlim=c(0,11),main="nu=0.01",xlab=xxlab1,las=2)
par(pop)
mtext("Number of Reticulations", outer = TRUE, cex = 1.5,side=3,line=-2)
```


### GLM Poison
```{r chunk5.1, cache = TRUE, echo=TRUE}
ModNu0.01<-glm(RetNum~CombHybP,data=outputNu0.01,family=poisson(link = "log"))
summary(ModNu0.01)

ModNu0.02<-glm(RetNum~CombHybP,data=outputNu0.02,family=poisson(link = "log"))
summary(ModNu0.02)

summary(ModNu0.01)
summary(ModNu0.02)
```


## Time

```{r chunk6, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10)}
pop<-par(mfrow=c(1,2),oma = c(0, 0, 2, 0))
vioplot(TimeTree~CombHybP,data=outputNu0.01,main="0.01",ylim=c(0,16),las=2)
vioplot(TimeTree~CombHybP,data=outputNu0.02,main="0.02",ylim=c(0,16),las=2)
par(pop)
mtext("Time to reach 15 tips", outer = TRUE, cex = 1.5,side=3,line=-2)

```


## Number of Nodes

```{r chunk7, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10)}
library(vioplot)
pop<-par(mfrow=c(1,2),oma = c(0, 0, 2, 0))
vioplot(NodeNum~CombHybP,data=outputNu0.01,main="nu = 0.01",ylim=c(14,33),las=2)
vioplot(NodeNum~CombHybP,data=outputNu0.02,main="nu = 0.02",ylim=c(14,33),las=2)
par(pop)
mtext("Number of Nodes", outer = TRUE, cex = 1.5,side=3,line=-2)
```



## Prop of Reticulations Not Recognozible

```{r chunk8, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10)}
## Prop of Reticulations Not Recognozible

SuboutputNu0.02Ret<-subset(outputNu0.02,RetNum>0)
tabwNu0.02Ret<-table(SuboutputNu0.02Ret[,c("NotIdent","CombHybP")])

tabw1Nu0.02Ret<-tabwNu0.02Ret
retTotNu0.02Ret<-apply(tabw1Nu0.02Ret,2,sum)

for(i in 1:ncol(tabw1Nu0.02Ret)){
tabw1Nu0.02Ret[,i]<-tabw1Nu0.02Ret[,i]/retTotNu0.02Ret[i]
}




SuboutputNu0.01Ret<-subset(outputNu0.01,RetNum>0)
tabwNu0.01Ret<-table(SuboutputNu0.01Ret[,c("NotIdent","CombHybP")])

tabw1Nu0.01Ret<-tabwNu0.01Ret
retTotNu0.01Ret<-apply(tabw1Nu0.01Ret,2,sum)

for(i in 1:ncol(tabw1Nu0.01Ret)){
tabw1Nu0.01Ret[,i]<-tabw1Nu0.01Ret[,i]/retTotNu0.01Ret[i]
}



pop<-par(mfrow=c(1,2),oma = c(0, 0, 2, 0))
barplot(tabw1Nu0.02Ret,legend.text=rownames(tabw1Nu0.02Ret),col=colors1,main="nu=0.02",xlim=c(0,11),xlab=xxlab1,las=2)
barplot(tabw1Nu0.01Ret,legend.text=rownames(tabw1Nu0.01Ret),col=colors1,main="nu=0.01",xlim=c(0,11),xlab=xxlab1,las=2)
par(pop)
mtext("Prop of Reticulations Not Recognozible", outer = TRUE, cex = 1.5,side=3,line=-2)


```



## Prop of Reticulations Not NotLevel.1

```{r chunk9, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10)}
## Prop of Reticulations Not NotLevel.1

###########
SuboutputNu0.02Ret<-subset(outputNu0.02,RetNum>0)
tabwNu0.02Ret<-table(SuboutputNu0.02Ret[,c("NotLevel.1","CombHybP")])

tabw1Nu0.02Ret<-tabwNu0.02Ret
retTotNu0.02Ret<-apply(tabw1Nu0.02Ret,2,sum)

for(i in 1:ncol(tabw1Nu0.02Ret)){
tabw1Nu0.02Ret[,i]<-tabw1Nu0.02Ret[,i]/retTotNu0.02Ret[i]
}




SuboutputNu0.01Ret<-subset(outputNu0.01,RetNum>0)
tabwNu0.01Ret<-table(SuboutputNu0.01Ret[,c("NotLevel.1","CombHybP")])

tabw1Nu0.01Ret<-tabwNu0.01Ret
retTotNu0.01Ret<-apply(tabw1Nu0.01Ret,2,sum)

for(i in 1:ncol(tabw1Nu0.01Ret)){
tabw1Nu0.01Ret[,i]<-tabw1Nu0.01Ret[,i]/retTotNu0.01Ret[i]
}
tabw1Nu0.01Ret


pop<-par(mfrow=c(1,2),oma = c(0, 0, 2, 0))
barplot(tabw1Nu0.02Ret,legend.text=rownames(tabw1Nu0.02Ret),col=colors1,main="nu=0.02",xlim=c(0,11),xlab=xxlab1,las=2)
barplot(tabw1Nu0.01Ret,legend.text=rownames(tabw1Nu0.01Ret),col=colors1,main="nu=0.01",xlim=c(0,11),xlab=xxlab1,las=2)
par(pop)
mtext("Prop of Reticulations Not NotLevel.1", outer = TRUE, cex = 1.5,side=3,line=-2)
```
