# Final Plots

## Reading in data

We are following (and copying) some of the lines from `results_summary.R`:

```{r}
library(tidyverse)
library(ggplot2)
library(survival)
library(xtable)
library(grafify)
library(patchwork)
library(ggridges)
library(vegan)
library(patchwork)

in_phy_filt <- read.csv('../summarized_results/rep_summed_filter_extended.csv')
#in_phy_filt <- read.csv('data-plots/rep_summed_filter_extended_2.csv')


## this is how we want to filter the results: by hmax=number of blobs, or hmax=1,2,3,4,5 or hmax=nrets
all_filts<-unique(in_phy_filt$filter)
```

We want first to do plots for hmax="blob"
```{r}
filter_type = "blob"

blob_phy_filt <- in_phy_filt %>%
    mutate(across(where(is.numeric), ~ if_else(.x < 0, NA, .x))) %>%
    filter(filter == filter_type) %>%
    mutate(
      # Convert grouping variables to factors so ggplot treats them as categories
      ntips_label = paste(ntips, "Tips"), # specific labels for the facets
      ngt = as.factor(ngt),
      level = if_else(level >= 5, "5≤", as.character(level)),
      level = factor(level, levels = c("1", "2", "3", "4", "5≤")),# bin
      # X-axis
    ) %>%
    ungroup()

```

# Main text figures

We are copying code from `make_plots.R` and modifying.

## Quantifying Topological Agreement to Ground Truth

### Quarnet Consistency Score

```{r}
cols <- c("Quarnet Precision" = "#1A5F7A", "Quarnet Recall" = "#D86C4E")
ggplot(blob_phy_filt, aes(x = level, group = ngt, linetype = ngt)) +
  # --- LEFT AXIS DATA 
  stat_summary(aes(y = quar_found, color = "Quarnet Precision"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = quar_found, color = "Quarnet Precision"), 
               fun = mean, geom = "point", size = 3) +
  # --- RIGHT AXIS DATA 
  stat_summary(aes(y = quar_compat, color = "Quarnet Recall"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = quar_compat, color = "Quarnet Recall"), 
               fun = mean, geom = "point", size = 3) +
  # --- DUAL AXIS SETUP ---
  scale_y_continuous(
    name = "Quarnet Precision",
    sec.axis = sec_axis(~ ., name = "Quarnet Recall")
  ) +
  facet_wrap(~ ntips_label) +
  scale_color_manual(values = cols) +
  guides(color = "none") + 
  theme_minimal() +
  theme(
    legend.key.width = unit(1, "cm"),
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold"),
    axis.title.y.left = element_text(color = cols["Quarnet Precision"], face = "bold"),
    axis.title.y.right = element_text(color = cols["Quarnet Recall"], face = "bold")
  ) +
  labs( #title = 'Quarnet Compatability Score',
        x = "Network Level",
        #color = "Metric",
        linetype = "Gene Trees")

ggsave('plots/quarnet_score_updated.png', width=8, height=4, units="in")
```

## Assessing Exact and Partial Recovery of Reticulations

### Recall and Precision

```{r}
cols <- c("Recall" = "#1A5F7A", "Precision" = "#D86C4E")
ltys <- c("100" = "solid", "1000" = "dashed", "10000" = "dotted")

ggplot(blob_phy_filt, aes(x = level, group = ngt, linetype = ngt)) +
  
  # --- LEFT AXIS DATA (RECALL / TPR) ---
  stat_summary(aes(y = tpr, color = "Recall"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = tpr, color = "Recall"), 
               fun = mean, geom = "point", size = 3) +
  # --- RIGHT AXIS DATA (PRECISION / PPV) ---
  stat_summary(aes(y = ppv, color = "Precision"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = ppv, color = "Precision"), 
               fun = mean, geom = "point", size = 3) +
  # --- DUAL AXIS SETUP ---
  scale_y_continuous(
    name = "Recall",
    sec.axis = sec_axis(~ ., name = "Precision")
  ) +
  facet_wrap(~ ntips_label) +
  scale_color_manual(values = cols) + ##manual color wrapping
  scale_linetype_manual(values = ltys) +
  guides(color = "none") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold"),
    axis.title.y.left = element_text(color = cols["Recall"], face = "bold"),
    axis.title.y.right = element_text(color = cols["Precision"], face = "bold")
  ) +
  labs(
    x = "Network Level",
    linetype = "Gene Trees")
ggsave('plots/recall_precision_hm_updated.png',width=8, height=4, units="in")
```

### Hybrid cluster broad coverage

```{r}
plot_dat <- blob_phy_filt %>%  group_by(ntips, level, ngt) %>%
  summarise(mean_pct = mean(broad_coverage, na.rm = TRUE) * 100, .groups = "drop") %>%
  mutate(
    ntips_label = paste(ntips, "Tips"),
    ngt = as.factor(ngt),
    level = as.factor(level)
  )

ggplot(plot_dat, aes(x = ngt, y = level, fill = mean_pct)) +
  geom_tile(color = "white", lwd = 0.5) +
  geom_text(aes(label = round(mean_pct, 1)), 
            color = "white", size = 3, fontface = "bold") +
  facet_wrap(~ ntips_label) +
  scale_fill_gradientn(colors = c("#FDE725","#21908C", "#440154"), 
                       name = "") +
  theme_minimal() +
  coord_fixed() +
  theme(
    panel.grid = element_blank(), # Heatmaps don't need gridlines
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "right"
  ) +
  labs(title = "Hybrid Clade Recall",
       x = "Gene Trees",
       y = "Network Level")

ggsave('plots/hyb_clade_broad_cov_hm_updated.png',width=6, height=4, units="in")
```

### Hybrid Cluster Broad Mapping

```{r}

plot_dat <- blob_phy_filt %>%  group_by(ntips, level, ngt) %>%
  summarise(mean_pct = mean(broad_mapping, na.rm = TRUE) * 100, .groups = "drop") %>%
  mutate(
    ntips_label = paste(ntips, "Tips"),
    ngt = as.factor(ngt),
    level = as.factor(level)
  )

ggplot(plot_dat, aes(x = ngt, y = level, fill = mean_pct)) +
  geom_tile(color = "white", lwd = 0.5) + # White borders make it look cleaner
  geom_text(aes(label = round(mean_pct, 1)), 
            color = "white", size = 3, fontface = "bold") +
  facet_wrap(~ ntips_label) +
  scale_fill_gradientn(colors = c("#FDE725","#21908C", "#440154"), 
                       name = "") +
  theme_minimal() +
  coord_fixed() + # Keeps the tiles square
  theme(
    panel.grid = element_blank(), # Heatmaps don't need gridlines
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "right"
  ) +
  labs(title = "Hybrid Clade Precision",
       x = "Gene Trees",
       y = "Network Level")

ggsave('plots/hyb_clade_broad_map_hm_updated.png',width=6, height=4, units="in")
```

## Evaluating Blob-Level Topological Agreement

### ToB Distances

```{r}



scale_factor <- 1
cols <- c("RF" = "#E69F00", "SplitInfo" = "#56B4E9")
ltys <- c("100" = "solid", "1000" = "dashed", "10000" = "dotted")
ggplot(blob_phy_filt, aes(x = level, group = ngt, linetype = ngt)) +
  # --- LEFT AXIS: RF DISTANCE (Primary) ---
  stat_summary(aes(y = rf_dist, color = "RF"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = rf_dist, color = "RF"), 
               fun = mean, geom = "point", size = 3) +
  # --- RIGHT AXIS: SPLIT INFO (Secondary - Scaled Down) ---
  # We DIVIDE by scale_factor here to bring the line down visually
  stat_summary(aes(y = split_dist / scale_factor, color = "SplitInfo"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = split_dist / scale_factor, color = "SplitInfo"), 
               fun = mean, geom = "point", size = 3) +
  # --- DUAL AXIS SETUP ---
  scale_y_continuous(
    name = "Robinson-Foulds Distance",
    # We MULTIPLY by scale_factor here to make the labels show the comparable numbers
    sec.axis = sec_axis(~ . * scale_factor, name = "Matching Split Information Distance")
  ) +
  facet_wrap(~ ntips_label) +
  scale_color_manual(values = cols) +
  scale_linetype_manual(values = ltys) +
  guides(color = "none") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(1, "cm"),
    strip.text = element_text(size = 12, face = "bold"),
    axis.title.y.left = element_text(color = cols["RF"], face = "bold"),
    axis.title.y.right = element_text(color = cols["SplitInfo"], face = "bold")
  ) +
  labs(x = "Network Level",
       linetype = "Gene Trees")

ggsave('plots/ToB_dists_updated.png',width=8, height=4, units="in")
```

### Blob Compatibility

```{r}

plot_dat <- blob_phy_filt %>%  group_by(ntips, level, ngt) %>%
  summarise(mean_pct = mean(perc_blobs_found, na.rm = TRUE) * 100, .groups = "drop") %>%
  mutate(
    ntips_label = paste(ntips, "Tips"),
    ngt = as.factor(ngt),
  )


ggplot(plot_dat, aes(x = ngt, y = level, fill = mean_pct)) +
  geom_tile(color = "white", lwd = 0.5) + # White borders make it look cleaner
  geom_text(aes(label = round(mean_pct, 1)), 
            color = "white", size = 3, fontface = "bold") +
  facet_wrap(~ ntips_label) +
  scale_fill_gradientn(colors = c("#FDE725","#21908C", "#440154"), 
                       name = "") +
  theme_minimal() +
  coord_fixed() + # Keeps the tiles square
  theme(
    panel.grid = element_blank(), # Heatmaps don't need gridlines
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "right"
  ) +
  labs(x = "Gene Trees",
       y = "Network Level")

ggsave('plots/tob_found_hm_updated.png',width=6, height=4, units="in")
```


### CF distances

```{r}

ggplot(blob_phy_filt, aes(x = level, y = CF_est_obs, fill = ngt)) +
  geom_violin(position = position_dodge(width = 0.8), 
              alpha = 0.6,          # Slightly more transparent to see gridlines
              trim = FALSE, 
              scale = "width",
              color = "grey50",     
              size = 0.3) +        
  #geom_boxplot(position = position_dodge(width = 0.8), 
  #             width = 0.15,        
  #             color = "grey20",    
  #             fill = "white",      
  #             alpha = 0.9,         
  #             outlier.shape = NA) +
  facet_wrap(~ ntips_label,ncol=1) +
  scale_fill_grafify(palette = "fishy") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(), # Cleaner background
    strip.text = element_text(size = 12, face = "bold"),
    axis.text = element_text(color = "grey30"),
    axis.title = element_text(face="bold",size=12),
    plot.title = element_text(face = "bold", size = 14)
  ) +
  labs(
    x = "Network Level",
    y = expression(italic(D)(italic(CF)[est], italic(CF)[obs])),
    fill = "Number of Gene Trees"
  )+
ylim(0,0.35)
ggsave('plots/CF_obs_est.png',width=4,height=6)
```

```{r}

ggplot(blob_phy_filt, aes(x = level, y = CF_obs_true, fill = ngt)) +
  geom_violin(position = position_dodge(width = 0.8), 
              alpha = 0.6,          # Slightly more transparent to see gridlines
              trim = FALSE, 
              scale = "width",
              color = "grey50",     
              size = 0.3) +        
  #geom_boxplot(position = position_dodge(width = 0.8), 
  #             width = 0.15,        
  #             color = "grey20",    
  #             fill = "white",      
  #             alpha = 0.9,         
  #             outlier.shape = NA) +
  facet_wrap(~ ntips_label,ncol=1) +
  scale_fill_grafify(palette = "fishy") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(), # Cleaner background
    strip.text = element_text(size = 12, face = "bold"),
    axis.text = element_text(color = "grey30"),
    axis.title = element_text(face="bold",size=12),
    plot.title = element_text(face = "bold", size = 14)
  ) +
  labs(
    x = "Network Level",
    y = expression(italic(D)(italic(CF)[true], italic(CF)[obs])),
    fill = "Number of Gene Trees"
  )

  ggsave('plots/CF_obs_true_dist.png',width = 4, height = 6)
```

```{r}
ggplot(blob_phy_filt, aes(x = level, y = CF_est_true, fill = ngt)) +
  geom_violin(position = position_dodge(width = 0.8), 
              alpha = 0.6,          # Slightly more transparent to see gridlines
              trim = FALSE, 
              scale = "width",
              color = "grey50",     
              size = 0.3) +        
  # geom_boxplot(position = position_dodge(width = 0.8), 
  #              width = 0.15,        
  #              color = "grey20",    
  #              fill = "white",      
  #              alpha = 0.9,         
  #              outlier.shape = NA) +
  facet_wrap(~ ntips_label,ncol=1) +
  scale_fill_grafify(palette = "fishy") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(), # Cleaner background
    strip.text = element_text(size = 12, face = "bold"),
    axis.text = element_text(color = "grey30"),
    axis.title = element_text(face="bold",size=12),
    plot.title = element_text(face = "bold", size = 14)
  ) +
  labs(
    x = "Network Level",
    y = expression(italic(D)(italic(CF)[est], italic(CF)[true])),
    fill = "Number of Gene Trees"
  )+
  ylim(0,0.35)
ggsave('plots/CF_est_true_dist.png',width = 4, height = 6)
```


```{r}
# ggplot(blob_phy_filt, aes(y = CF_obs_true, x = CF_est_obs,color=ngt)) +
#   geom_point(alpha = 0.4, size = 1.5) +
#   # Adds the 1:1 diagonal line
#   geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40", size = 0.8) +
#   scale_color_grafify(palette = "fishy") +
#   theme_minimal() +
#   theme(
#     legend.position = "bottom",
#     panel.grid.minor = element_blank(),
#     axis.text = element_text(color = "grey30"),
#     axis.title = element_text(face="bold",size=12),
#     plot.title = element_text(face = "bold", size = 14)
#   ) +
#   labs(
#     #title = "Estimation Error vs. Sampling Noise",
#     y = expression(italic(D)(italic(CF)[true], italic(CF)[obs])) ,
#     x = expression(italic(D)(italic(CF)[est], italic(CF)[obs])),
#     color = "Number of Gene Trees",
#     shape = "number of Tips"
#   ) +
#   # Optional: keep axes square to make the diagonal meaningful
#   ylim(0, 0.125) + 
#   xlim(0, 0.4)
# ggsave('plots/CF_scatter_noise_obs_fit.png',height=4,width=4)
```

```{r}
library(ggplot2)
library(patchwork)
library(grafify)

# Generalized plotting function
make_scatter_plot <- function(df, 
                              x_var, 
                              y_var, 
                              ngt_val, 
                              x_lim = NULL, 
                              y_lim = NULL, 
                              x_label = NULL, 
                              y_label = NULL, 
                              add_diagonal = TRUE) {
  
  # Look up panel color by ngt value
  point_color <- custom_colors[as.character(ngt_val)]
  
  # Filter data
  plot_data <- dplyr::filter(df, ngt == ngt_val)
  
  # Base plot using .data[[var]] for flexible column selection
  p <- ggplot(plot_data, aes(x = .data[[x_var]], y = .data[[y_var]]))
  
  # Optional 1:1 diagonal line
  if (add_diagonal) {
    p <- p + geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.8)
  }
  
  p <- p +
    geom_point(color = point_color, alpha = 0.5, size = 1.5) +
    scale_x_continuous(limits = x_lim) +
    scale_y_continuous(limits = y_lim) +
    labs(
      title = paste(ngt_val, "Gene Trees"),
      x = x_label,
      y = y_label
    ) +
    theme_minimal() +
    theme(
      legend.position = "none",
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 12, hjust = 0.5, color = point_color),
      axis.text = element_text(color = "grey30"),
      axis.title = element_text(face = "bold", size = 11)
    )
  
  return(p)
}


# Optional: Extract specific colors from the 'fishy' palette (or provide custom hex codes)
custom_colors <- c(
  "100"   = "#90B0D0",  # Muted steel blue
  "1000"  = "#FBBF69",  # Soft yellow/amber
  "10000" = "#F59896"   # Muted salmon/coral
)
# 2. Build the three plots passing distinct colors
x_expr <- expression(italic(D)(italic(CF)[est], italic(CF)[obs]))
y_expr <- expression(italic(D)(italic(CF)[true], italic(CF)[obs]))

p100   <- make_scatter_plot(blob_phy_filt, "CF_est_obs", "CF_obs_true", 100,   x_lim = c(0, 0.25), y_lim = c(0.04, 0.12), x_label = x_expr, y_label = y_expr)
p1000  <- make_scatter_plot(blob_phy_filt, "CF_est_obs", "CF_obs_true", 1000,  x_lim = c(0, 0.20), y_lim = c(0.01, 0.05), x_label = x_expr, y_label = NULL)
p10000 <- make_scatter_plot(blob_phy_filt, "CF_est_obs", "CF_obs_true", 10000, x_lim = c(0, 0.15), y_lim = c(0.00, 0.02), x_label = x_expr, y_label = NULL)
# 3. Stitch together
combined_plot <- p100 + p1000 + p10000
combined_plot
# Adjust width so the panels aren't squished horizontally
ggsave('plots/CF_scatter_noise_obs_fit.png', height = 4, width = 10)

```

```{r}
# 
# ggplot(blob_phy_filt, aes(y = CF_obs_true, x = CF_est_true,color=ngt)) +
#   geom_point(alpha = 0.4, size = 1.5) +
#   # Adds the 1:1 diagonal line
#   geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40", size = 0.8) +
#   scale_color_grafify(palette = "fishy") +
#   theme_minimal() +
#   theme(
#     legend.position = "bottom",
#     panel.grid.minor = element_blank(),
#     axis.text = element_text(color = "grey30"),
#     axis.title = element_text(face="bold",size=12),
#     plot.title = element_text(face = "bold", size = 14)
#   ) +
#   labs(
#     #title = "Estimation Error vs. Sampling Noise",
#    y = expression(italic(D)(italic(CF)[true], italic(CF)[obs])),
#     x = expression(italic(D)(italic(CF)[est], italic(CF)[true])),
#     color = "Number of Gene Trees",
#     shape = "Number of Tips"
#   ) +
#   # Optional: keep axes square to make the diagonal meaningful
#   ylim(0, 0.125) + 
#   xlim(0, 0.4)


p100   <- make_scatter_plot(blob_phy_filt, "CF_est_true", "CF_obs_true", 100,   x_lim = c(0, 0.25), y_lim = c(0.04, 0.12), x_label = x_expr, y_label = y_expr)
p1000  <- make_scatter_plot(blob_phy_filt, "CF_est_true", "CF_obs_true", 1000,  x_lim = c(0, 0.20), y_lim = c(0.01, 0.05), x_label = x_expr, y_label = NULL)
p10000 <- make_scatter_plot(blob_phy_filt, "CF_est_true", "CF_obs_true", 10000, x_lim = c(0, 0.15), y_lim = c(0.00, 0.02), x_label = x_expr, y_label = NULL)
# 3. Stitch together
combined_plot <- p100 + p1000 + p10000
combined_plot

ggsave('plots/CF_scatter_noise_true_error.png',height=4,width=10)
```




# Supplemental figures

## Filtered by 'blob'

### Hardwired cluster distance
```{r}
ggplot(blob_phy_filt, aes(x = level, y = hwcd, fill = ngt)) +
  geom_violin(position = position_dodge(width = 0.8), 
              alpha = 0.6,          # Slightly more transparent to see gridlines
              trim = FALSE, 
              scale = "width",
              color = "grey50",     
              size = 0.3) +        
  geom_boxplot(position = position_dodge(width = 0.8), 
               width = 0.15,        
               color = "grey20",    
               #fill = "white",      
               alpha = 0.9,         
               outlier.shape = NA) +
  facet_wrap(~ ntips_label) +
  scale_fill_grafify(palette = "fishy") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(), # Cleaner background
    strip.text = element_text(size = 12, face = "bold"),
    axis.text = element_text(color = "grey30"),
    plot.title = element_text(face = "bold", size = 14)
  ) +
  labs(
    x = "Network Level",
    y = 'Hardwired Cluster Dissimilarity',
    fill = "Gene Trees"
  )

ggsave('plots/HWCD_updated.png',width=8, height=4, units="in")
```

### Exact coverage

```{r}
plot_dat <- blob_phy_filt %>%  group_by(ntips, level, ngt) %>%
  summarise(mean_pct = mean(exact_coverage, na.rm = TRUE) * 100, .groups = "drop") %>%
  mutate(
    ntips_label = paste(ntips, "Tips"),
    ngt = as.factor(ngt),
    level = as.factor(level)
  )


ggplot(plot_dat, aes(x = ngt, y = level, fill = mean_pct)) +
  geom_tile(color = "white", lwd = 0.5) + # White borders make it look cleaner
  geom_text(aes(label = round(mean_pct, 1)), 
            color = "white", size = 3, fontface = "bold") +

  facet_wrap(~ ntips_label) +
  scale_fill_gradientn(colors = c("#FDE725", "#21908C", "#440154"), 
                       name = "") +
  
  theme_minimal() +
  coord_fixed() + # Keeps the tiles square
  theme(
    panel.grid = element_blank(), # Heatmaps don't need gridlines
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "right"
  ) +
  labs(x = "Gene Trees",
       y = "Network Level")

ggsave('plots/hyb_clade_ex_cov_hm_updated.png',width=6, height=4, units="in")
```

### Compatible blobs

```{r}
plot_dat <- blob_phy_filt %>%  group_by(ntips, level, ngt) %>%
  summarise(mean_pct = mean(perc_blob_compat, na.rm = TRUE) * 100, .groups = "drop") %>%
  mutate(
    ntips_label = paste(ntips, "Tips"),
    ngt = as.factor(ngt),
  )

ggplot(plot_dat, aes(x = ngt, y = level, fill = mean_pct)) +
  geom_tile(color = "white", lwd = 0.5) + # White borders make it look cleaner
  geom_text(aes(label = round(mean_pct, 1)), 
            color = "white", size = 3, fontface = "bold") +
  facet_wrap(~ ntips_label) +
  scale_fill_gradientn(colors = c("#FDE725", "#21908C", "#440154"), 
                       name = "") +
  theme_minimal() +
  coord_fixed() + # Keeps the tiles square
  theme(
    panel.grid = element_blank(), # Heatmaps don't need gridlines
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "right"
  ) +
  labs(x = "Gene Trees",
       y = "Network Level")

ggsave('plots/tob_compat_hm_updated.png',width=6, height=4, units="in")
```

### Reduced distances

```{r}
cols <- c("Canonical HWCD" = "#1A5F7A", "FU-Stable HWCD" = "#D86C4E")
ltys <- c("100" = "solid", "1000" = "dashed", "10000" = "dotted")

ggplot(blob_phy_filt, aes(x = level, group = ngt, linetype = ngt)) +
  # --- LEFT AXIS DATA 
  stat_summary(aes(y = canon_hwcd, color = "Canonical HWCD"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = canon_hwcd, color = "Canonical HWCD"), 
               fun = mean, geom = "point", size = 3) +
  # --- RIGHT AXIS DATA 
  stat_summary(aes(y = fu_hwcd, color = "FU-Stable HWCD"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = fu_hwcd, color = "FU-Stable HWCD"), 
               fun = mean, geom = "point", size = 3) +
  # --- DUAL AXIS SETUP ---
  scale_y_continuous(
    name = "Canonical HWCD",
    sec.axis = sec_axis(~ ., name = "FU-Stable HWCD")
  ) +
  facet_wrap(~ ntips_label) +
  scale_color_manual(values = cols) +
  scale_linetype_manual(values = ltys) +
  scale_linetype_manual(values = ltys) +
  guides(color = "none") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(1, "cm"),
    strip.text = element_text(size = 12, face = "bold"),
    axis.title.y.left = element_text(color = cols["Canonical HWCD"], face = "bold"),
    axis.title.y.right = element_text(color = cols["FU-Stable HWCD"], face = "bold")
  ) +
  labs(
    x = "Network Level",
    linetype = "Gene Trees")

ggsave('plots/reduced_dists_updated.png',width=8, height=4, units="in")
```

### Displayed HWCD

```{r}
# scale_factor <- 0.1
# cols <- c("total" = "#E69F00", "average" = "#56B4E9")
# 
# ggplot(blob_phy_filt, aes(x = level, group = ngt, linetype = ngt)) +
#   stat_summary(aes(y = total_disp_rf, color = "total"), 
#                fun = mean, geom = "line", size = 1.2) +
#   stat_summary(aes(y = total_disp_rf, color = "total"), 
#                fun = mean, geom = "point", size = 3) +
#   stat_summary(aes(y = ave_disp_rf / scale_factor, color = "average"), 
#                fun = mean, geom = "line", size = 1.2) +
#   stat_summary(aes(y = ave_disp_rf / scale_factor, color = "average"), 
#                fun = mean, geom = "point", size = 3) +
#   # --- DUAL AXIS SETUP ---
#   scale_y_continuous(
#     name = "Total paired RF distance",
#     # We MULTIPLY by scale_factor here to make the labels show the comparable numbers
#     sec.axis = sec_axis(~ . * scale_factor, name = "Average paired RF distance")
#   ) +
#   facet_wrap(~ ntips_label) +
#   scale_color_manual(values = cols) +
#   guides(color = "none") +
#   theme_minimal() +
#   theme(
#     legend.position = "bottom",
#     strip.text = element_text(size = 12, face = "bold"),
#     # Match Axis Colors for Clarity
#     axis.title.y.left = element_text(color = cols["total"], face = "bold"),
#     axis.title.y.right = element_text(color = cols["average"], face = "bold")
#   ) +
#   labs(x = "Network Level",
#        linetype = "Gene Trees")
# 
# ggsave('plots/displayed_hwcd_updated.png',width=8, height=4, units="in")

##Make two plots 
# 1. Plot for Total Paired RF Distance
ltys <- c("100" = "solid", "1000" = "dashed", "10000" = "dotted")

p_total <- ggplot(blob_phy_filt, aes(x = level, group = ngt, linetype = ngt)) +
  stat_summary(aes(y = total_disp_rf), fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = total_disp_rf), fun = mean, geom = "point", size = 3) +
  scale_y_continuous(name = "Total paired RF distance") +
  facet_wrap(~ ntips_label) +
  scale_linetype_manual(values = ltys) +
  theme_minimal() +
  theme(legend.position = "bottom", axis.title.y = element_text(face = "bold"),
legend.key.width = unit(1, "cm")) +
  labs(x = "Network Level", linetype = "Gene Trees", color = NULL)

# 2. Plot for Average Paired RF Distance
p_ave <- ggplot(blob_phy_filt, aes(x = level, group = ngt, linetype = ngt)) +
  stat_summary(aes(y = ave_disp_rf), fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = ave_disp_rf), fun = mean, geom = "point", size = 3) +
  scale_y_continuous(name = "Average paired RF distance") +
  facet_wrap(~ ntips_label) +
  scale_linetype_manual(values = ltys) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.title.y = element_text(face = "bold"),
        legend.key.width = unit(1, "cm")) +
  labs(x = "Network Level", linetype = "Gene Trees", color = NULL)

p_total
p_ave
# Save both plots
ggsave('plots/total_disp_rf.png', p_total, width=8, height=4, units="in")
ggsave('plots/average_disp_rf.png', p_ave, width=8, height=4, units="in")


```

## Filtered by 'nrets'

```{r}
filter_type<-'nrets'

blob_phy_filt <- in_phy_filt %>%
    mutate(across(where(is.numeric), ~ if_else(.x < 0, NA, .x))) %>%
    filter(filter == filter_type) %>%
    mutate(
      # Convert grouping variables to factors so ggplot treats them as categories
      ntips_label = paste(ntips, "Tips"), # specific labels for the facets
      ngt = as.factor(ngt),
      level = if_else(level >= 5, "5≤", as.character(level)),
      level = factor(level, levels = c("1", "2", "3", "4", "5≤")),# bin
      # X-axis
    ) %>%
    ungroup()
```

### HWCD

```{r}
ggplot(blob_phy_filt, aes(x = level, y = hwcd, fill = ngt)) +
  geom_violin(position = position_dodge(width = 0.8), 
              alpha = 0.6,          # Slightly more transparent to see gridlines
              trim = FALSE, 
              scale = "width",
              color = "grey50",     
              size = 0.3) +        
  geom_boxplot(position = position_dodge(width = 0.8), 
               width = 0.15,        
               color = "grey20",    
               #fill = "white",      
               alpha = 0.9,         
               outlier.shape = NA) +
  facet_wrap(~ ntips_label) +
  scale_fill_grafify(palette = "fishy") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(), # Cleaner background
    strip.text = element_text(size = 12, face = "bold"),
    axis.text = element_text(color = "grey30"),
    plot.title = element_text(face = "bold", size = 14)
  ) +
  labs(
    x = "Network Level",
    y = 'Hardwired Cluster Dissimilarity',
    fill = "Gene Trees"
  )

ggsave('plots/HWCD_nrets_updated.png',width=8, height=4, units="in")
```

### Recall-precision

```{r}
cols <- c("Recall" = "#1A5F7A", "Precision" = "#D86C4E")
ltys <- c("100" = "solid", "1000" = "dashed", "10000" = "dotted")
ggplot(blob_phy_filt, aes(x = level, group = ngt, linetype = ngt)) +
  # --- LEFT AXIS DATA (RECALL / TPR) ---
  stat_summary(aes(y = tpr, color = "Recall"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = tpr, color = "Recall"), 
               fun = mean, geom = "point", size = 3) +
  # --- RIGHT AXIS DATA (PRECISION / PPV) ---
  stat_summary(aes(y = ppv, color = "Precision"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = ppv, color = "Precision"), 
               fun = mean, geom = "point", size = 3) +
  # --- DUAL AXIS SETUP ---
  scale_y_continuous(
    name = "Recall",
    sec.axis = sec_axis(~ ., name = "Precision")
  ) +
  facet_wrap(~ ntips_label) +
  scale_color_manual(values = cols) + ##manual color wrapping
scale_linetype_manual(values = ltys) +
  guides(color = "none") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(1, "cm"),
    strip.text = element_text(size = 12, face = "bold"),
    axis.title.y.left = element_text(color = cols["Recall"], face = "bold"),
    axis.title.y.right = element_text(color = cols["Precision"], face = "bold")
  ) +
  labs(
    x = "Network Level",
    linetype = "Gene Trees")

ggsave('plots/recall_precision_hm_nrets_updated.png',width=8, height=4, units="in")
```

### Found blobs

```{r}
plot_dat <- blob_phy_filt %>%  group_by(ntips, level, ngt) %>%
  summarise(mean_pct = mean(perc_blobs_found, na.rm = TRUE) * 100, .groups = "drop") %>%
  mutate(
    ntips_label = paste(ntips, "Tips"),
    ngt = as.factor(ngt),
  )


ggplot(plot_dat, aes(x = ngt, y = level, fill = mean_pct)) +
  geom_tile(color = "white", lwd = 0.5) + # White borders make it look cleaner
  geom_text(aes(label = round(mean_pct, 1)), 
            color = "white", size = 3, fontface = "bold") +
  facet_wrap(~ ntips_label) +
  scale_fill_gradientn(colors = c("#FDE725", "#21908C", "#440154"), 
                       name = "") +
  theme_minimal() +
  coord_fixed() + # Keeps the tiles square
  theme(
    panel.grid = element_blank(), # Heatmaps don't need gridlines
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "right"
  ) +
  labs(x = "Gene Trees",
       y = "Network Level")

ggsave('plots/tob_found_hm_nrets_updated.png',width=6, height=4, units="in")
```

