# Final Plots

From Josh's message:
I just pushed to the `granular_pipeline` branch! The main file is in the `pipeline` folder of `level_2`.
Plots are mostly made out of `filtered_sum.R` which does the following:

- summarize and plot simulation-only information (L18-101)
  - with the exception of the branch length data, you should be able to replicate this with `sim_net_props.csv`

- summarize analysis results (L115-195)
  - This section produces the aggregated results data across simulation settings and hmax values (averaged across replicates). The CSV with this info is `rep_summed_filter_extended.csv`

- Make the plots outlined in excalidraw, for each hmax setting (1-5, number of blobs and number of reticulations; L206-235) 
  - The core plotting is actually done in the script `make_plots.R`

- Make plots but filter by the number of genes and summarize across hmax settings (L269-806)

New message by Josh:
Ah sorry! Not `filtered_sum.R`, it should be `results_summary.R`.
`blob_phy_filt` gets generated in the for loop of of L211 of this file.

It uses the `rep_summed_filter_extended.csv` to generate the dataset for each different way we could filter the data and runs `make_plots.R` to make the same plots with the filter.

I placed `sim_net_props.csv` and `rep_summed_filter_extended.csv` inside a `data-plots` folder (in `pipeline`), but git ignored.

## Reading in data

We are following (and copying) some of the lines from `results_summary.R`:

```{r}
library(tidyverse)
library(ggplot2)
library(survival)
library(xtable)
library(grafify)
library(patchwork)
library(ggridges)
library(vegan)

in_phy_filt <- read.csv('data-plots/rep_summed_filter_extended.csv')

## this is how we want to filter the results: by hmax=number of blobs, or hmax=1,2,3,4,5 or hmax=nrets
all_filts<-unique(in_phy_filt$filter)
```

We want first to do plots for hmax="blob"
```{r}
filter_type = "blob"

blob_phy_filt <- in_phy_filt %>%
    mutate(across(where(is.numeric), ~ if_else(.x < 0, NA, .x))) %>%
    filter(filter == filter_type) %>%
    mutate(
      # Convert grouping variables to factors so ggplot treats them as categories
      ntips_label = paste(ntips, "Tips"), # specific labels for the facets
      ngt = as.factor(ngt),
      level = if_else(level >= 5, "5≤", as.character(level)),
      level = factor(level, levels = c("1", "2", "3", "4", "5≤")),# bin
      # X-axis
    ) %>%
    ungroup()

```

# Main text figures

We are copying code from `make_plots.R` and modifying.

## Quantifying Topological Agreement to Ground Truth

### Quarnet Consistency Score

```{r}
cols <- c("Found Quarnets" = "#1A5F7A", "Compatible Quarnets" = "#D86C4E")
ggplot(blob_phy_filt, aes(x = level, group = ngt, linetype = ngt)) +
  # --- LEFT AXIS DATA 
  stat_summary(aes(y = quar_found, color = "Found Quarnets"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = quar_found, color = "Found Quarnets"), 
               fun = mean, geom = "point", size = 3) +
  # --- RIGHT AXIS DATA 
  stat_summary(aes(y = quar_compat, color = "Compatible Quarnets"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = quar_compat, color = "Compatible Quarnets"), 
               fun = mean, geom = "point", size = 3) +
  # --- DUAL AXIS SETUP ---
  scale_y_continuous(
    name = "Found Quarnets",
    sec.axis = sec_axis(~ ., name = "Compatible Quarnets")
  ) +
  facet_wrap(~ ntips_label) +
  scale_color_manual(values = cols) +
  guides(color = "none") + 
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold"),
    axis.title.y.left = element_text(color = cols["Found Quarnets"], face = "bold"),
    axis.title.y.right = element_text(color = cols["Compatible Quarnets"], face = "bold")
  ) +
  labs( #title = 'Quarnet Compatability Score',
        x = "Network Level",
        #color = "Metric",
        linetype = "Gene Trees")

ggsave('plots/quarnet_score_updated.png', width=8, height=4, units="in")
```

## Assessing Exact and Partial Recovery of Reticulations

### Recall and Precision

```{r}
cols <- c("Recall" = "#1A5F7A", "Precision" = "#D86C4E")
ggplot(blob_phy_filt, aes(x = level, group = ngt, linetype = ngt)) +
  
  # --- LEFT AXIS DATA (RECALL / TPR) ---
  stat_summary(aes(y = tpr, color = "Recall"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = tpr, color = "Recall"), 
               fun = mean, geom = "point", size = 3) +
  # --- RIGHT AXIS DATA (PRECISION / PPV) ---
  stat_summary(aes(y = ppv, color = "Precision"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = ppv, color = "Precision"), 
               fun = mean, geom = "point", size = 3) +
  # --- DUAL AXIS SETUP ---
  scale_y_continuous(
    name = "Recall",
    sec.axis = sec_axis(~ ., name = "Precision")
  ) +
  facet_wrap(~ ntips_label) +
  scale_color_manual(values = cols) + ##manual color wrapping
  guides(color = "none") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold"),
    axis.title.y.left = element_text(color = cols["Recall"], face = "bold"),
    axis.title.y.right = element_text(color = cols["Precision"], face = "bold")
  ) +
  labs(
    x = "Network Level",
    linetype = "Gene Trees")
ggsave('plots/recall_precision_hm_updated.png',width=8, height=4, units="in")
```

### Hybrid cluster broad coverage

```{r}
plot_dat <- blob_phy_filt %>%  group_by(ntips, level, ngt) %>%
  summarise(mean_pct = mean(broad_coverage, na.rm = TRUE) * 100, .groups = "drop") %>%
  mutate(
    ntips_label = paste(ntips, "Tips"),
    ngt = as.factor(ngt),
    level = as.factor(level)
  )

ggplot(plot_dat, aes(x = ngt, y = level, fill = mean_pct)) +
  geom_tile(color = "white", lwd = 0.5) +
  geom_text(aes(label = round(mean_pct, 1)), 
            color = "white", size = 3, fontface = "bold") +
  facet_wrap(~ ntips_label) +
  scale_fill_gradientn(colors = c("#FDE725","#21908C", "#440154"), 
                       name = "") +
  theme_minimal() +
  coord_fixed() +
  theme(
    panel.grid = element_blank(), # Heatmaps don't need gridlines
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "right"
  ) +
  labs(title = "Found True Hybrid",
       x = "Gene Trees",
       y = "Network Level")

ggsave('plots/hyb_clade_broad_cov_hm_updated.png',width=4, height=4, units="in")
```

### Hybrid Cluster Broad Mapping

```{r}

plot_dat <- blob_phy_filt %>%  group_by(ntips, level, ngt) %>%
  summarise(mean_pct = mean(broad_mapping, na.rm = TRUE) * 100, .groups = "drop") %>%
  mutate(
    ntips_label = paste(ntips, "Tips"),
    ngt = as.factor(ngt),
    level = as.factor(level)
  )

ggplot(plot_dat, aes(x = ngt, y = level, fill = mean_pct)) +
  geom_tile(color = "white", lwd = 0.5) + # White borders make it look cleaner
  geom_text(aes(label = round(mean_pct, 1)), 
            color = "white", size = 3, fontface = "bold") +
  facet_wrap(~ ntips_label) +
  scale_fill_gradientn(colors = c("#FDE725","#21908C", "#440154"), 
                       name = "") +
  theme_minimal() +
  coord_fixed() + # Keeps the tiles square
  theme(
    panel.grid = element_blank(), # Heatmaps don't need gridlines
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "right"
  ) +
  labs(title = "Compatible Estimated Hybrid",
       x = "Gene Trees",
       y = "Network Level")

ggsave('plots/hyb_clade_broad_map_hm_updated.png',width=4, height=4, units="in")
```

## Evaluating Blob-Level Topological Agreement

### ToB Distances

```{r}

scale_factor <- 1
cols <- c("RF" = "#E69F00", "SplitInfo" = "#56B4E9")

ggplot(blob_phy_filt, aes(x = level, group = ngt, linetype = ngt)) +
  # --- LEFT AXIS: RF DISTANCE (Primary) ---
  stat_summary(aes(y = rf_dist, color = "RF"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = rf_dist, color = "RF"), 
               fun = mean, geom = "point", size = 3) +
  # --- RIGHT AXIS: SPLIT INFO (Secondary - Scaled Down) ---
  # We DIVIDE by scale_factor here to bring the line down visually
  stat_summary(aes(y = split_dist / scale_factor, color = "SplitInfo"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = split_dist / scale_factor, color = "SplitInfo"), 
               fun = mean, geom = "point", size = 3) +
  # --- DUAL AXIS SETUP ---
  scale_y_continuous(
    name = "Robinson-Foulds Distance",
    # We MULTIPLY by scale_factor here to make the labels show the comparable numbers
    sec.axis = sec_axis(~ . * scale_factor, name = "Split Information Distance")
  ) +
  facet_wrap(~ ntips_label) +
  scale_color_manual(values = cols) +
  guides(color = "none") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold"),
    axis.title.y.left = element_text(color = cols["RF"], face = "bold"),
    axis.title.y.right = element_text(color = cols["SplitInfo"], face = "bold")
  ) +
  labs(x = "Network Level",
       linetype = "Gene Trees")

ggsave('plots/ToB_dists_updated.png',width=8, height=4, units="in")
```

### Blob Compatibility

```{r}

plot_dat <- blob_phy_filt %>%  group_by(ntips, level, ngt) %>%
  summarise(mean_pct = mean(perc_blobs_found, na.rm = TRUE) * 100, .groups = "drop") %>%
  mutate(
    ntips_label = paste(ntips, "Tips"),
    ngt = as.factor(ngt),
  )


ggplot(plot_dat, aes(x = ngt, y = level, fill = mean_pct)) +
  geom_tile(color = "white", lwd = 0.5) + # White borders make it look cleaner
  geom_text(aes(label = round(mean_pct, 1)), 
            color = "white", size = 3, fontface = "bold") +
  facet_wrap(~ ntips_label) +
  scale_fill_gradientn(colors = c("#FDE725","#21908C", "#440154"), 
                       name = "") +
  theme_minimal() +
  coord_fixed() + # Keeps the tiles square
  theme(
    panel.grid = element_blank(), # Heatmaps don't need gridlines
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "right"
  ) +
  labs(x = "Gene Trees",
       y = "Network Level")

ggsave('plots/tob_found_hm_updated.png',width=4, height=4, units="in")
```

# Supplemental figures

## Filtered by 'blob'

### Hardwired cluster distance
```{r}
ggplot(blob_phy_filt, aes(x = level, y = hwcd, fill = ngt)) +
  geom_violin(position = position_dodge(width = 0.8), 
              alpha = 0.6,          # Slightly more transparent to see gridlines
              trim = FALSE, 
              scale = "width",
              color = "grey50",     
              size = 0.3) +        
  geom_boxplot(position = position_dodge(width = 0.8), 
               width = 0.15,        
               color = "grey20",    
               fill = "white",      
               alpha = 0.9,         
               outlier.shape = NA) +
  facet_wrap(~ ntips_label) +
  scale_fill_grafify(palette = "fishy") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(), # Cleaner background
    strip.text = element_text(size = 12, face = "bold"),
    axis.text = element_text(color = "grey30"),
    plot.title = element_text(face = "bold", size = 14)
  ) +
  labs(
    x = "Network Level",
    y = 'Hardwired Cluster Distance',
    fill = "Gene Trees"
  )

ggsave('plots/HWCD_updated.png',width=8, height=4, units="in")
```

### Exact coverage

```{r}
plot_dat <- blob_phy_filt %>%  group_by(ntips, level, ngt) %>%
  summarise(mean_pct = mean(exact_coverage, na.rm = TRUE) * 100, .groups = "drop") %>%
  mutate(
    ntips_label = paste(ntips, "Tips"),
    ngt = as.factor(ngt),
    level = as.factor(level)
  )


ggplot(plot_dat, aes(x = ngt, y = level, fill = mean_pct)) +
  geom_tile(color = "white", lwd = 0.5) + # White borders make it look cleaner
  geom_text(aes(label = round(mean_pct, 1)), 
            color = "white", size = 3, fontface = "bold") +

  facet_wrap(~ ntips_label) +
  scale_fill_gradientn(colors = c("#FDE725", "#21908C", "#440154"), 
                       name = "") +
  
  theme_minimal() +
  coord_fixed() + # Keeps the tiles square
  theme(
    panel.grid = element_blank(), # Heatmaps don't need gridlines
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "right"
  ) +
  labs(x = "Gene Trees",
       y = "Network Level")

ggsave('plots/hyb_clade_ex_cov_hm_updated.png',width=4, height=4, units="in")
```

### Compatible blobs

```{r}
plot_dat <- blob_phy_filt %>%  group_by(ntips, level, ngt) %>%
  summarise(mean_pct = mean(perc_blob_compat, na.rm = TRUE) * 100, .groups = "drop") %>%
  mutate(
    ntips_label = paste(ntips, "Tips"),
    ngt = as.factor(ngt),
  )

ggplot(plot_dat, aes(x = ngt, y = level, fill = mean_pct)) +
  geom_tile(color = "white", lwd = 0.5) + # White borders make it look cleaner
  geom_text(aes(label = round(mean_pct, 1)), 
            color = "white", size = 3, fontface = "bold") +
  facet_wrap(~ ntips_label) +
  scale_fill_gradientn(colors = c("#FDE725", "#21908C", "#440154"), 
                       name = "") +
  theme_minimal() +
  coord_fixed() + # Keeps the tiles square
  theme(
    panel.grid = element_blank(), # Heatmaps don't need gridlines
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "right"
  ) +
  labs(x = "Gene Trees",
       y = "Network Level")

ggsave('plots/tob_compat_hm_updated.png',width=4, height=4, units="in")
```

### Reduced distances

```{r}
cols <- c("Canonical HWCD" = "#1A5F7A", "FU-Stable HWCD" = "#D86C4E")
ggplot(blob_phy_filt, aes(x = level, group = ngt, linetype = ngt)) +
  # --- LEFT AXIS DATA 
  stat_summary(aes(y = canon_hwcd, color = "Canonical HWCD"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = canon_hwcd, color = "Canonical HWCD"), 
               fun = mean, geom = "point", size = 3) +
  # --- RIGHT AXIS DATA 
  stat_summary(aes(y = fu_hwcd, color = "FU-Stable HWCD"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = fu_hwcd, color = "FU-Stable HWCD"), 
               fun = mean, geom = "point", size = 3) +
  # --- DUAL AXIS SETUP ---
  scale_y_continuous(
    name = "Canonical HWCD",
    sec.axis = sec_axis(~ ., name = "FU-Stable HWCD")
  ) +
  facet_wrap(~ ntips_label) +
  scale_color_manual(values = cols) +
  guides(color = "none") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold"),
    axis.title.y.left = element_text(color = cols["Canonical HWCD"], face = "bold"),
    axis.title.y.right = element_text(color = cols["FU-Stable HWCD"], face = "bold")
  ) +
  labs(
    x = "Network Level",
    linetype = "Gene Trees")

ggsave('plots/reduced_dists_updated.png',width=8, height=4, units="in")
```

### Displayed HWCD

```{r}
scale_factor <- 0.1
cols <- c("total" = "#E69F00", "average" = "#56B4E9")

ggplot(blob_phy_filt, aes(x = level, group = ngt, linetype = ngt)) +
  stat_summary(aes(y = total_disp_rf, color = "total"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = total_disp_rf, color = "total"), 
               fun = mean, geom = "point", size = 3) +
  stat_summary(aes(y = ave_disp_rf / scale_factor, color = "average"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = ave_disp_rf / scale_factor, color = "average"), 
               fun = mean, geom = "point", size = 3) +
  # --- DUAL AXIS SETUP ---
  scale_y_continuous(
    name = "Total paired RF distance",
    # We MULTIPLY by scale_factor here to make the labels show the comparable numbers
    sec.axis = sec_axis(~ . * scale_factor, name = "Average paired RF distance")
  ) +
  facet_wrap(~ ntips_label) +
  scale_color_manual(values = cols) +
  guides(color = "none") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold"),
    # Match Axis Colors for Clarity
    axis.title.y.left = element_text(color = cols["total"], face = "bold"),
    axis.title.y.right = element_text(color = cols["average"], face = "bold")
  ) +
  labs(x = "Network Level",
       linetype = "Gene Trees")

ggsave('plots/displayed_hwcd_updated.png',width=8, height=4, units="in")
```

## Filtered by 'nrets'

```{r}
filter_type<-'nrets'

blob_phy_filt <- in_phy_filt %>%
    mutate(across(where(is.numeric), ~ if_else(.x < 0, NA, .x))) %>%
    filter(filter == filter_type) %>%
    mutate(
      # Convert grouping variables to factors so ggplot treats them as categories
      ntips_label = paste(ntips, "Tips"), # specific labels for the facets
      ngt = as.factor(ngt),
      level = if_else(level >= 5, "5≤", as.character(level)),
      level = factor(level, levels = c("1", "2", "3", "4", "5≤")),# bin
      # X-axis
    ) %>%
    ungroup()
```

### HWCD

```{r}
ggplot(blob_phy_filt, aes(x = level, y = hwcd, fill = ngt)) +
  geom_violin(position = position_dodge(width = 0.8), 
              alpha = 0.6,          # Slightly more transparent to see gridlines
              trim = FALSE, 
              scale = "width",
              color = "grey50",     
              size = 0.3) +        
  geom_boxplot(position = position_dodge(width = 0.8), 
               width = 0.15,        
               color = "grey20",    
               fill = "white",      
               alpha = 0.9,         
               outlier.shape = NA) +
  facet_wrap(~ ntips_label) +
  scale_fill_grafify(palette = "fishy") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(), # Cleaner background
    strip.text = element_text(size = 12, face = "bold"),
    axis.text = element_text(color = "grey30"),
    plot.title = element_text(face = "bold", size = 14)
  ) +
  labs(
    x = "Network Level",
    y = 'Hardwired Cluster Distance',
    fill = "Gene Trees"
  )

ggsave('plots/HWCD_nrets_updated.png',width=8, height=4, units="in")
```

### Recall-precision

```{r}
cols <- c("Recall" = "#1A5F7A", "Precision" = "#D86C4E")
ggplot(blob_phy_filt, aes(x = level, group = ngt, linetype = ngt)) +
  # --- LEFT AXIS DATA (RECALL / TPR) ---
  stat_summary(aes(y = tpr, color = "Recall"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = tpr, color = "Recall"), 
               fun = mean, geom = "point", size = 3) +
  # --- RIGHT AXIS DATA (PRECISION / PPV) ---
  stat_summary(aes(y = ppv, color = "Precision"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = ppv, color = "Precision"), 
               fun = mean, geom = "point", size = 3) +
  # --- DUAL AXIS SETUP ---
  scale_y_continuous(
    name = "Recall",
    sec.axis = sec_axis(~ ., name = "Precision")
  ) +
  facet_wrap(~ ntips_label) +
  scale_color_manual(values = cols) + ##manual color wrapping
  guides(color = "none") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold"),
    axis.title.y.left = element_text(color = cols["Recall"], face = "bold"),
    axis.title.y.right = element_text(color = cols["Precision"], face = "bold")
  ) +
  labs(
    x = "Network Level",
    linetype = "Gene Trees")

ggsave('plots/recall_precision_hm_nrets_updated.png',width=8, height=4, units="in")
```

### Found blobs

```{r}
plot_dat <- blob_phy_filt %>%  group_by(ntips, level, ngt) %>%
  summarise(mean_pct = mean(perc_blobs_found, na.rm = TRUE) * 100, .groups = "drop") %>%
  mutate(
    ntips_label = paste(ntips, "Tips"),
    ngt = as.factor(ngt),
  )


ggplot(plot_dat, aes(x = ngt, y = level, fill = mean_pct)) +
  geom_tile(color = "white", lwd = 0.5) + # White borders make it look cleaner
  geom_text(aes(label = round(mean_pct, 1)), 
            color = "white", size = 3, fontface = "bold") +
  facet_wrap(~ ntips_label) +
  scale_fill_gradientn(colors = c("#FDE725", "#21908C", "#440154"), 
                       name = "") +
  theme_minimal() +
  coord_fixed() + # Keeps the tiles square
  theme(
    panel.grid = element_blank(), # Heatmaps don't need gridlines
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "right"
  ) +
  labs(x = "Gene Trees",
       y = "Network Level")

ggsave('plots/tob_found_hm_nrets_updated.png',width=4, height=4, units="in")
```