# Final Plots

From Josh's message:
I just pushed to the `granular_pipeline` branch! The main file is in the `pipeline` folder of `level_2`.
Plots are mostly made out of `filtered_sum.R` which does the following:

- summarize and plot simulation-only information (L18-101)
  - with the exception of the branch length data, you should be able to replicate this with `sim_net_props.csv`

- summarize analysis results (L115-195)
  - This section produces the aggregated results data across simulation settings and hmax values (averaged across replicates). The CSV with this info is `rep_summed_filter_extended.csv`

- Make the plots outlined in excalidraw, for each hmax setting (1-5, number of blobs and number of reticulations; L206-235) 
  - The core plotting is actually done in the script `make_plots.R`

- Make plots but filter by the number of genes and summarize across hmax settings (L269-806)

New message by Josh:
Ah sorry! Not `filtered_sum.R`, it should be `results_summary.R`.
`blob_phy_filt` gets generated in the for loop of of L211 of this file.

It uses the `rep_summed_filter_extended.csv` to generate the dataset for each different way we could filter the data and runs `make_plots.R` to make the same plots with the filter.


I placed `sim_net_props.csv` and `rep_summed_filter_extended.csv` inside a `data-plots` folder (in `pipeline`), but git ignored.

## Reading in data

We are following (and copying) some of the lines from `results_summary.R`:

```{r}
library(tidyverse)
library(ggplot2)
library(survival)
library(xtable)
library(grafify)
library(patchwork)
library(ggridges)
library(vegan)

in_phy_filt <- read.csv('data-plots/rep_summed_filter_extended.csv')

## this is how we want to filter the results: by hmax=number of blobs, or hmax=1,2,3,4,5 or hmax=nrets
all_filts<-unique(in_phy_filt$filter)
```

We want first to do plots for hmax="blob"
```{r}
filter_type = "blob"

blob_phy_filt <- in_phy_filt %>%
    mutate(across(where(is.numeric), ~ if_else(.x < 0, NA, .x))) %>%
    filter(filter == filter_type) %>%
    mutate(
      # Convert grouping variables to factors so ggplot treats them as categories
      ntips_label = paste(ntips, "Tips"), # specific labels for the facets
      ngt = as.factor(ngt),
      level = if_else(level >= 5, "5≤", as.character(level)),
      level = factor(level, levels = c("1", "2", "3", "4", "5≤")),# bin
      # X-axis
    ) %>%
    ungroup()

```

# Main text figures

We are copying code from `make_plots.R` and modifying.

## Quarnet Consistency Score

```{r}
cols <- c("Found Quarnets" = "#1A5F7A", "Compatible Quarnets" = "#D86C4E")
ggplot(blob_phy_filt, aes(x = level, group = ngt, linetype = ngt)) +
  
  # --- LEFT AXIS DATA 
  stat_summary(aes(y = quar_found, color = "Found Quarnets"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = quar_found, color = "Found Quarnets"), 
               fun = mean, geom = "point", size = 3) +
  # --- RIGHT AXIS DATA 
  stat_summary(aes(y = quar_compat, color = "Compatible Quarnets"), 
               fun = mean, geom = "line", size = 1.2) +
  stat_summary(aes(y = quar_compat, color = "Compatible Quarnets"), 
               fun = mean, geom = "point", size = 3) +
  # --- DUAL AXIS SETUP ---
  scale_y_continuous(
    name = "Found Quarnets",
    sec.axis = sec_axis(~ ., name = "Compatible Quarnets")
  ) +
  facet_wrap(~ ntips_label) +
  scale_color_manual(values = cols) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold"),
    axis.title.y.left = element_text(color = cols["Found Quarnets"], face = "bold"),
    axis.title.y.right = element_text(color = cols["Compatible Quarnets"], face = "bold")
  ) +
  labs( title = 'Quarnet Compatability Score',
        x = "Network Level",
        color = "Metric",
        linetype = "Gene Trees")

ggsave(paste(save_dir,'quarnet_score.png',sep=''))
```

## Evaluating the Accuracy of Inferred Reticulations
