---
title: "Verifying binary trees"
author: 
    - Gustavo A. Ballen^[Museu de Zoologia da Universidade de São Paulo]
output: 
    pdf_document:
        fig_caption: yes
        number_sections: true
urlcolor: blue
toc: true
---

```{r setup, echo = FALSE}
#library(kableExtra)
library(knitr)
knit_hooks$set(small.mar = function(before, options, envir) {
    if (before)    par(mar=c(5,4,1,1)+.1)  # smaller margin on top and right
})
knitr::opts_chunk$set(tidy = TRUE, small.mar = TRUE, comment = "")
```

```{r}
library(NetSim) # simulate bd time-tree networks: https://github.com/jjustison/NetSim
library(TreeSim) # simulate bd time-trees
library(ape) # tools for handling phylo objects
library(diversitree) # tools for estimating lambda and mu
library(lattice) # stacked plots
library(phytools)
set.seed(1000)
numbsim <- 100
```

Check in which situations the `.age` functions in `TreeSim` and `NetSim` produce binary trees. The whole thing should be controlled through the argument `mrca` in both packages. Also, it is possible that malformed trees come from short-lived processes where the whole tree becomes extinct or just fails to speciate; this can be explored using three arbitrary levels of time: 4, 8, and 10 time units.

In addition, the function `ape::is.binary` tests whether the tree is binary so the whole trick is to test with mrca = TRUE and FALSE for each packages (i.e., 12 simulations), then apply the function `ape::is.binary` to each tree in a given collection and summarize the number of binary trees (if any).

```{r}
mrca <- c(TRUE, FALSE)
age <- c(4, 6, 8)
simNames <- apply(X = expand.grid(mrca,
                                  age),
                  MARGIN = 1,
                  FUN = function(x) paste(x, collapse = "-"))
simTreeSim <- list()
length(simTreeSim) <- length(simNames)
#names(simTreeSim) <- simNames
simNetSim <- list()
length(simNetSim) <- length(simNames)
#names(simNetSim) <- simNames
```

Begin simulating trees for `TreeSim`:

```{r, cache = TRUE}
# using TreeSim
nIter <- 1
for (i in mrca) {
    for (j in age) {
        cat("Creating sim with TreeSim, ", "mrca = ", i, " age = ", j, "\n", sep = "")
        simTreeSim[nIter] <- list(eval(call("sim.bd.age",
                                       age = j,
                                       numbsim = numbsim,
                                       lambda = 1,
                                       mu = 0,
                                       frac = 1,
                                       mrca = FALSE,
                                       complete = TRUE,
                                       K = 0)))
        simNames[nIter] <- paste("TS", i, j, sep = "_")
        print(simNames[nIter])
        nIter <- nIter + 1
    }
}
names(simTreeSim) <- simNames
```

Now simulating trees for `NetSim`:

```{r, cache = TRUE}
# using NetSim
nIter <- 1
for (i in mrca) {
    for (j in age) {
        cat("Creating sim with NetSim, ", "mrca = ", i, " age = ", j, "\n", sep = "")
        simNetSim[[nIter]] <- eval(call("sim.bdh.age",
                                      age = j,
                                      numbsim = numbsim,
                                      lambda = 1,
                                      mu = 0,
                                      nu = 0,
                                      hybprops = c(0.5, 0.5, 0.5),
                                      hyb.inher.fxn = NetSim::make.beta.draw(1,1),
                                      frac = 1,
                                      mrca = i,
                                      complete = TRUE,
                                      stochsampling = FALSE,
                                      hyb.rate.fxn = NULL,
                                      trait.model = NULL))
        simNames[nIter] <- paste("NS", i, j, sep = "_")
        print(simNames[nIter])
        nIter <- nIter + 1
    }
}
names(simNetSim) <- simNames
```

Summarize binary trees for `TreeSim` in each experiment:

```{r}
# create an object for results of iterative testing whether trees are binary
bins_TreeSim <- list()
length(bins_TreeSim) <- length(simTreeSim)
index <- 1

# change the class for each experiment list to multiPhylo
simTreeSim <- lapply(simTreeSim, FUN = function(x) `class<-`(x, "multiPhylo"))

for (i in simTreeSim) {
    # remove elements that are not of class phylo
    i <- i[sapply(X = i, FUN = function(x) inherits(x, "phylo"))]
    # remove elements that are null
    i <- i[!sapply(X = i, FUN = is.null)]
    # append the freq table of binary to non-binary trees to output object
    bins_TreeSim[index] <- list(table(is.binary(i)))
    # update index
    index <- index + 1
}

# update names and show results for TreeSim
# please note that numbers other than 100 represent cases where
# some simulated elements were either non-phylo or null
names(bins_TreeSim) <- names(simTreeSim)
print(bins_TreeSim)
```

Summarize binary trees for `NetSim` in each experiment:

```{r}
# create an object for results of iterative testing whether trees are binary
bins_NetSim <- list()
length(bins_NetSim) <- length(simNetSim)
index <- 1

simNetSim <- lapply(simNetSim, FUN = function(x) `class<-`(x, "multiPhylo"))

for (i in simNetSim) {
    # remove elements that are not of class phylo
    i <- i[sapply(X = i, FUN = function(x) inherits(x, "phylo"))]
    # remove elements that are null
    i <- i[!sapply(X = i, FUN = is.null)]
    # append the freq table of binary to non-binary trees to output object
    bins_NetSim[index] <- list(table(is.binary(i)))
    # update index
    index <- index + 1
}

# update names and show results for NetSim
# please note that numbers other than 100 represent cases where
# some simulated elements were either non-phylo or null
names(bins_NetSim) <- names(simNetSim)
print(bins_NetSim)
```
