---
title: "Assessing bias features in measures of center in Yule trees"
author: 
    - Gustavo A. Ballen^[Museu de Zoologia da Universidade de São Paulo]
output: 
    pdf_document:
        fig_caption: yes
        number_sections: true
urlcolor: blue
toc: true
---

```{r setup, echo = FALSE}
#library(kableExtra)
library(knitr)
knit_hooks$set(small.mar = function(before, options, envir) {
    if (before)    par(mar=c(5,4,1,1)+.1)  # smaller margin on top and right
})
knitr::opts_chunk$set(tidy = TRUE, small.mar = TRUE, comment = "")
```

Start with TreeSim

```{r}
library(NetSim) # simulate bd time-tree networks: https://github.com/jjustison/NetSim
library(TreeSim) # simulate bd time-trees
library(ape) # tools for handling phylo objects
library(diversitree) # tools for estimating lambda and mu
#library(lattice) # stacked plots
library(phytools)
set.seed(1000)
```

Let's explore the relationship between convergence (or the lack thereof, due to the initial value in the mle), type of center measure, and magnitude in bias usin Carlos' initial code in an iterative way for automatic experiments. All simulations are using the .age functions with Yule trees

Using TreeSim:

```{r, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10)}
# based on the original code of bias calculation by Carlos
# iterating with an approach similar to binary-property-simulations.R
numbsim <- 100
x.init <- c(0.1, 0.5, 0.9)
center <- c("mean", "median")

# Yule, TreeSim, age = 4, mrca = T
par(mfrow = c(3, 2), oma=c(1,1,1,1), mar=c(2,2,2,2))
for (a in x.init) {
    for (b in center) {
        biasLamTS<-c()
        for(i in 1:100){
            yuleTS <- TreeSim::sim.bd.age(age = 4,
                                 numbsim = numbsim,
                                 lambda = 1,
                                 mu = 0,
                                 frac = 1,
                                 mrca = TRUE,
                                 complete = TRUE,
                                 K = 0)
            ## remove any trees with no taxa
            yuleTS <- yuleTS[!sapply(X = yuleTS, FUN = is.null)]
            ## remove non-phylo elements
            yuleTS <- yuleTS[sapply(X = yuleTS, FUN = is.phylo)]
            ## remove any trees with <10 taxa
            yuleTS <- yuleTS[sapply(X = yuleTS, FUN = function(x) length(x$tip.label) >= 10)]
            length(yuleTS)
            # fit a Yule model with diversitree
            yuleTSFits <- lapply(X = yuleTS, FUN = function(x) diversitree::make.yule(tree = x,sampling.f = 1))
            # fetch the coefficients estimates on the previous lik funs
            yuleTSMLEcoefs <- lapply(X = yuleTSFits, FUN = function(x) coef(diversitree::find.mle(func = x,x.init = a)))
            #Lambda Bias
            biasYule <- 1 - eval(call(b, unlist(yuleTSMLEcoefs)))
            biasLamTS <- c(biasLamTS, biasYule)
        }
        hist(biasLamTS, main = paste("x.init = ", a, ", estimator = ", b, "\n", sep = ""))
    }
}
```

Now using NetSim:

```{r, cache = TRUE, fig.asp = 1.0, dev.args  = list(pointsize = 10)}
# based on the original code of bias calculation by Carlos
# iterating with an approach similar to binary-property-simulations.R
numbsim <- 100
x.init <- c(0.1, 0.5, 0.9)
center <- c("mean", "median")

# Yule, TreeSim, age = 4, mrca = T
par(mfrow = c(3, 2), oma=c(1,1,1,1), mar=c(2,2,2,2))
for (a in x.init) {
    for (b in center) {
        biasLamNS<-c()
        for(i in 1:100){
            yuleNS <- NetSim::sim.bdh.age(age = 4,
                                          numbsim = numbsim,
                                          lambda = 1,
                                          mu = 0,
                                          nu = 0,
                                          hybprops = c(0.5, 0.5, 0.5),
                                          hyb.inher.fxn = NetSim::make.beta.draw(1, 1),
                                          frac = 1,
                                          mrca = TRUE,
                                          complete = TRUE,
                                          stochsampling = FALSE,
                                          hyb.rate.fxn = NULL,
                                          trait.model = NULL)
            ## remove any trees with no taxa
            yuleNS <- yuleNS[!sapply(X = yuleNS, FUN = is.null)]
            ## remove non-phylo elements
            yuleNS <- yuleNS[sapply(X = yuleNS, FUN = is.phylo)]
            ## remove any trees with <10 taxa
            yuleNS <- yuleNS[sapply(X = yuleNS, FUN = function(x) length(x$tip.label) >= 10)]
            length(yuleNS)
            # fit a Yule model with diversitree
            yuleNSFits <- lapply(X = yuleNS, FUN = function(x) diversitree::make.yule(tree = x,sampling.f = 1))
            # fetch the coefficients estimates on the previous lik funs
            yuleNSMLEcoefs <- lapply(X = yuleNSFits, FUN = function(x) coef(diversitree::find.mle(func = x,x.init = a)))
            #Lambda Bias
            biasYule <- 1 - eval(call(b, unlist(yuleNSMLEcoefs)))#Bias
            biasLamNS <- c(biasLamNS, biasYule)
        }
        hist(biasLamNS, main = paste("x.init = ", a, ", estimator = ", b, "\n", sep = ""))
    }
}
```

There seems to be a similar behavior between packages. Both consistently recover lower mean and median values with respect to the theoretical value of $\lambda = 1.0$. The fact that the distribution of medians seems a bit leaned towards 0.0 suggests that the distribution of the mean is a bit asymmetric. In almost all simulations the highest value of bias is 0.06, so that it is probably tolerable. 

It seems irrelevant, but does the fact that we are calculating bias as 1.0[i.e., the real value of lambda] - mean(sample of lambdas) introduce any issue such as rounding bias or something else? A small modification should be made so that instead of 1.0 - mean(lambdas) we use mean([1.0 - lambdas)? It should not, for 1 - mean(lambdas) is algebraically == mean(1 - lambdas) being lambdas a vector of lambdas estimated from a sample o trees. The code was actually run before realizing the equivalence between both approaches and is included below just for sake of completeness.

Using TreeSim:

```{r, cache = TRUE, eval = FALSE, fig.asp = 1.0, dev.args  = list(pointsize = 10)}
# based on the original code of bias calculation by Carlos
# iterating with an approach similar to binary-property-simulations.R
numbsim <- 100
x.init <- c(0.1, 0.5, 0.9)
center <- c("mean", "median")

# Yule, TreeSim, age = 4, mrca = T
par(mfrow = c(3, 2), oma=c(1,1,1,1), mar=c(2,2,2,2))
for (a in x.init) {
    for (b in center) {
        biasLamTS<-c()
        for(i in 1:100){
            yuleTS <- TreeSim::sim.bd.age(age = 4,
                                 numbsim = numbsim,
                                 lambda = 1,
                                 mu = 0,
                                 frac = 1,
                                 mrca = TRUE,
                                 complete = TRUE,
                                 K = 0)
            ## remove any trees with no taxa
            yuleTS <- yuleTS[!sapply(X = yuleTS, FUN = is.null)]
            ## remove non-phylo elements
            yuleTS <- yuleTS[sapply(X = yuleTS, FUN = is.phylo)]
            ## remove any trees with <10 taxa
            yuleTS <- yuleTS[sapply(X = yuleTS, FUN = function(x) length(x$tip.label) >= 10)]
            length(yuleTS)
            # fit a Yule model with diversitree
            yuleTSFits <- lapply(X = yuleTS, FUN = function(x) diversitree::make.yule(tree = x,sampling.f = 1))
            # fetch the coefficients estimates on the previous lik funs
            yuleTSMLEcoefs <- lapply(X = yuleTSFits, FUN = function(x) coef(diversitree::find.mle(func = x,x.init = a)))
            #Lambda Bias
            biasYule <- eval(call(b, 1 - unlist(yuleTSMLEcoefs)))
            biasLamTS <- c(biasLamTS, biasYule)
        }
        hist(biasLamTS, main = paste("x.init = ", a, ", estimator = ", b, "\n", sep = ""))
    }
}
```

Now using NetSim:

```{r, cache = TRUE, eval = FALSE, fig.asp = 1.0, dev.args  = list(pointsize = 10)}
# based on the original code of bias calculation by Carlos
# iterating with an approach similar to binary-property-simulations.R
numbsim <- 100
x.init <- c(0.1, 0.5, 0.9)
center <- c("mean", "median")

# Yule, TreeSim, age = 4, mrca = T
par(mfrow = c(3, 2), oma=c(1,1,1,1), mar=c(2,2,2,2))
for (a in x.init) {
    for (b in center) {
        biasLamNS<-c()
        for(i in 1:100){
            yuleNS <- NetSim::sim.bdh.age(age = 4,
                                          numbsim = numbsim,
                                          lambda = 1,
                                          mu = 0,
                                          nu = 0,
                                          hybprops = c(0.5, 0.5, 0.5),
                                          hyb.inher.fxn = NetSim::make.beta.draw(1, 1),
                                          frac = 1,
                                          mrca = TRUE,
                                          complete = TRUE,
                                          stochsampling = FALSE,
                                          hyb.rate.fxn = NULL,
                                          trait.model = NULL)
            ## remove any trees with no taxa
            yuleNS <- yuleNS[!sapply(X = yuleNS, FUN = is.null)]
            ## remove non-phylo elements
            yuleNS <- yuleNS[sapply(X = yuleNS, FUN = is.phylo)]
            ## remove any trees with <10 taxa
            yuleNS <- yuleNS[sapply(X = yuleNS, FUN = function(x) length(x$tip.label) >= 10)]
            length(yuleNS)
            # fit a Yule model with diversitree
            yuleNSFits <- lapply(X = yuleNS, FUN = function(x) diversitree::make.yule(tree = x,sampling.f = 1))
            # fetch the coefficients estimates on the previous lik funs
            yuleNSMLEcoefs <- lapply(X = yuleNSFits, FUN = function(x) coef(diversitree::find.mle(func = x,x.init = a)))
            #Lambda Bias
            biasYule <- eval(call(b, 1.0 - unlist(yuleNSMLEcoefs)))
            biasLamNS <- c(biasLamNS, biasYule)
        }
        hist(biasLamNS, main = paste("x.init = ", a, ", estimator = ", b, "\n", sep = ""))
    }
}
```

The last plots did not differ too much from the earlier ones, so mean(1 - lambda) vs. 1 - mean(lambda) (being lambda a vector) seems to be equivalent
