---
title: "Overview of the `NetSim` package"
author: 
    - Gustavo A. Ballen^[Museu de Zoologia da Universidade de São Paulo]
output: 
    pdf_document:
        fig_caption: yes
        number_sections: true
        citation_package: natbib
        keep_tex: true
        keep_md: true
urlcolor: blue
toc: true
biblio-style: apalike
bibliography: /home/balleng/Dropbox/Gustavo/Bibtex/PhDThesis.bib
header-includes: 
- \usepackage{lscape}
- \usepackage{longtable}
---

```{r setup, echo = FALSE}
#library(kableExtra)
library(knitr)
knit_hooks$set(small.mar = function(before, options, envir) {
    if (before)    par(mar=c(5,4,1,1)+.1)  # smaller margin on top and right
})
knitr::opts_chunk$set(tidy = TRUE, small.mar = TRUE, comment = "")
```

```{r}
library(NetSim) # simulate bd time-tree networks: https://github.com/jjustison/NetSim
library(TreeSim) # simulate bd time-trees
library(ape) # tools for handling phylo objects
library(diversitree) # tools for estimating lambda and mu
library(lattice) # stacked plots
set.seed(1000)
numbsim <- 100
```

# The NetSim package

The package builds on the approach of Stadler for `TreeSim`, but with modifications by Zhang et al. (2018) for including hybridization as part of the birth-death (BD) stochastic process. The functions are named similarly to those in `TreeSim` but with `bdh` insted of just `bd`. 

The package allows trees conditional to time or number of species, as well as the typical sampling and extinction patterns allowed by `TreeSim`. However, it allows to model hybridization in three important ways:

- Effect of the hybridization on generation of lineages (positive, negative, neutral)
- Hybridization probability as a function of genetic distance (the closer two lineages the easier for them to hybridize)
- Hybridization probability covariating with a trait along the phylogeny (for example, if hybridization is dependent on the two parents having a similar trait)

Simulations are controlled via two main parameters: Origin time, and number of species. For the former, a given time interval is specified for the simulation to run, leaving $n$ as a free quantity. On the other hand, the SSE and GSE approaches fix the final number of species $n$ and allow the simulation to run an unconstrained amount of time until such quantity is attained.

# General network simulation features

Comparing the manual entry for the functions `sim.bdh.age`, `sim.bdh.taxa.ssa`, and `sim.bdh.taxa.gsa`, we find that they differ in the very first arguments (Table \ref{tab:sim_args}). These first arguments (`age`, `m`, and `n`) control the endpoint of the simulation. All of the remaining arguments specify the model oh birth-death hybridization. All functions return an object of either class `phylo` or `multiphylo` depending on how many trees were requested in the argument `numbsim`. 

Below we will explore each of these functions and their appropriate context.

# Preliminary conditions: Yule and BD trees

Let's see if the functions produce reliable trees under Yule and BD without hybridization. Parameter estimation on the tree samples is inspired in this blog post by [Revell and Harmon](http://www.phytools.org/Cordoba2017/ex/12/BD-models.html)

## Yule trees

First let's simulate the same tree samples using `TreeSim` as a control for comparing the results by `NetSim`:

```{r, cache = TRUE}
yuleTS <- sim.bd.age(age = 4,
                     numbsim = numbsim,
                     lambda = 1,
                     mu = 0,
                     frac = 1,
                     mrca = FALSE,
                     complete = TRUE,
                     K = 0)

## remove any trees with no taxa
yuleTS <- yuleTS[!sapply(X = yuleTS, FUN = is.null)]
## remove non-phylo elements
yuleTS <- yuleTS[sapply(X = yuleTS, FUN = is.phylo)] 
## remove any trees with <10 taxa
yuleTS <- yuleTS[sapply(X = yuleTS, FUN = function(x) length(x$tip.label) >= 10)]
# fit a Yule model with diversitree
yuleTSFits <- lapply(X = yuleTS,
                FUN = function(x) diversitree::make.yule(tree = x,
                                                         sampling.f = 1.0))
# fetch the coefficients estimates on the previous lik funs
yuleTSMLEcoefs <- lapply(X = yuleTSFits,
                   FUN = function(x) coef(diversitree::find.mle(func = x,
                                                                x.init = 0.5)))
```

Now we will simulate a Yule tree sample using `NetSim`:

```{r, cache = TRUE}
# age = 4 should generate moderate-sized trees
# mu = 0 no extinction
# nu = 0 no hybridization
# hybprobs = c(0.5, 0.5, 0.5) no prior difference among types of effects of hyb on descendant lineages
# hyb.inher.fxn beta(1,1) -> uniform(0,1), no pattern of inheritance,
# frac = 1 we'll simulate complete genealogies
# complete = TRUE we don't have extinction, this plays no role
# hyb.rate.fxn = NULL we don't want hybridization to be a function of genetic distance
# trait.model = NULL we don't need a covariate to include in this simple model
yuleNS <- NetSim::sim.bdh.age(age = 4,
                         numbsim = numbsim,
                         lambda = 1,
                         mu = 0,
                         nu = 0,
                         hybprops = c(0.5, 0.5, 0.5),
                         hyb.inher.fxn = NetSim::make.beta.draw(1,1),
                         frac = 1, 
                         mrca = TRUE,
                         complete = TRUE, 
                         stochsampling = FALSE,
                         hyb.rate.fxn = NULL, 
                         trait.model = NULL 
                         )

## remove any trees with no taxa
yuleNS <- yuleNS[!sapply(X = yuleNS, FUN = is.null)] 
## remove non-phylo elements
yuleNS <- yuleNS[sapply(X = yuleNS, FUN = is.phylo)] 
## remove any trees with <10 taxa
yuleNS <- yuleNS[sapply(X = yuleNS, FUN = function(x) length(x$tip.label) >= 10)]
# fit a Yule model with diversitree
yuleNSFits <- lapply(X = yuleNS,
                FUN = function(x) diversitree::make.yule(tree = x,
                                                         sampling.f = 1.0))
# fetch the coefficients estimates on the previous lik funs
yuleNSMLEcoefs <- lapply(X = yuleNSFits,
                   FUN = function(x) coef(diversitree::find.mle(func = x,
                                                                x.init = 0.5)))
```

Plot both distributions for visual comparison. Please note that the original value of $\lambda$ used in the simulations is represented by the vertical black line:

```{r}
yuleLambda <- data.frame(lambda = c(unlist(yuleTSMLEcoefs), unlist(yuleNSMLEcoefs)),
                         method = c(rep("TreeSim", times = length(yuleTSMLEcoefs)),
                                    rep("NetSim", times = length(yuleNSMLEcoefs))))
yuleLambdaHist <- lattice::densityplot(~ lambda | method,
                                     data = yuleLambda,
                                     layout = c(1, 2),
                                     plot.points = FALSE,
                                     panel = function(x, y, ...) {
                                       panel.densityplot(x, ...)
                                       panel.abline(v=1.0)
                                     })
my_theme <- trellis.par.get()
my_theme$strip.background$col <- "grey80"
my_theme$plot.symbol$col <- "grey60"
my_theme$plot.polygon$col <- "grey90"

yuleLHplot <- update(yuleLambdaHist, par.settings = my_theme)
print(yuleLHplot)

```

## Birth-death trees

```{r, cache = TRUE}
# mrca = FALSE because the tree needs to be binary. This param behaves the opposite in NetSim!
# complete = FALSE drops extinct tips b/c diversitree needs an ultrametric tree
bdTS <- sim.bd.age(age = 9,
                     numbsim = numbsim,
                     lambda = 0.9,
                     mu = 0.5,
                     frac = 0.9,
                     mrca = FALSE,
                     complete = FALSE,
                     K = 0)

## remove any trees with no taxa
bdTS <- bdTS[!sapply(X = bdTS, FUN = is.null)]
## remove non-phylo elements
bdTS <- bdTS[sapply(X = bdTS, FUN = is.phylo)] 
## remove any trees with <10 taxa
bdTS <- bdTS[sapply(X = bdTS, FUN = function(x) length(x$tip.label) >= 10)]
# fit a Bd model with diversitree
bdTSFits <- lapply(X = bdTS,
                FUN = function(x) diversitree::make.bd(tree = x,
                                                         sampling.f = 0.6))
# fetch the coefficients estimates on the previous lik funs
bdTSMLEcoefs <- lapply(X = bdTSFits,
                   FUN = function(x) coef(diversitree::find.mle(func = x,
                                                                x.init = c(0.5, 0.5))))
```

Now we will simulate a BD tree sample using `NetSim`:

```{r, cache = TRUE}
# age = 9 should generate moderate-sized trees
# nu = 0 no hybridization
# hybprobs = c(0.5, 0.5, 0.5) no prior difference among types of effects of hyb on descendant lineages
# hyb.inher.fxn beta(1,1) -> uniform(0,1), no pattern of inheritance,
# frac = 1 we'll simulate complete genealogies
# mrca = TRUE because the tree needs to be binary
# complete = FALSE drop extinct tips b/c diversitree needs an ultrametric tree
# hyb.rate.fxn = NULL we don't want hybridization to be a function of genetic distance
# trait.model = NULL we don't need a covariate to include in this simple model
bdNS <- NetSim::sim.bdh.age(age = 9,
                         numbsim = numbsim,
                         lambda = 0.9,
                         mu = 0.5,
                         nu = 0,
                         hybprops = c(0.5, 0.5, 0.5),
                         hyb.inher.fxn = NetSim::make.beta.draw(1,1),
                         frac = 0.9, 
                         mrca = TRUE,
                         complete = FALSE, 
                         stochsampling = FALSE,
                         hyb.rate.fxn = NULL, 
                         trait.model = NULL 
                         )

## remove any trees with no taxa
bdNS <- bdNS[!sapply(X = bdNS, FUN = is.null)] 
## remove non-phylo elements
bdNS <- bdNS[sapply(X = bdNS, FUN = is.phylo)] 
## remove any trees with <10 taxa
bdNS <- bdNS[sapply(X = bdNS, FUN = function(x) length(x$tip.label) >= 10)]
# remove non-binary
bdNS <- bdNS[sapply(X = bdNS, FUN = is.binary)] 
# fit a Bd model with diversitree
bdNSFits <- lapply(X = bdNS,
                FUN = function(x) diversitree::make.bd(tree = x,
                                                         sampling.f = 0.6))
# fetch the coefficients estimates on the previous lik funs
bdNSMLEcoefs <- lapply(X = bdNSFits,
                   FUN = function(x) coef(diversitree::find.mle(func = x,
                                                                x.init = c(0.5, 0.5))))
```

Plot both distributions for visual comparisonPlease note that the original value of $\lambda$ and $\mu$ used in the simulations is represented by the vertical black line:

```{r}
bdParams <- data.frame(lambda = c(sapply(bdTSMLEcoefs, function(x) unlist(x)["lambda"]),
                                  sapply(bdNSMLEcoefs, function(x) unlist(x)["lambda"])),
                       mu = c(sapply(bdTSMLEcoefs, function(x) unlist(x)["mu"]),
                                  sapply(bdNSMLEcoefs, function(x) unlist(x)["mu"])), 
                         method = c(rep("TreeSim", times = length(bdTSMLEcoefs)),
                                    rep("NetSim", times = length(bdNSMLEcoefs))))

# build plot
bdLambdaHist <- lattice::densityplot(~ lambda | method,
                                     data = bdParams,
                                     layout = c(1, 2),
                                     plot.points = FALSE, 
                                     panel = function(x, y, ...) {
                                       panel.densityplot(x, ...)
                                       panel.abline(v=0.9)
                                     })
my_theme <- trellis.par.get()
my_theme$strip.background$col <- "grey80"
my_theme$plot.symbol$col <- "grey60"
my_theme$plot.polygon$col <- "grey90"

bdLHplot <- update(bdLambdaHist, par.settings = my_theme)
print(bdLHplot)

bdMuHist <- lattice::densityplot(~ mu | method,
                                     data = bdParams,
                                     layout = c(1, 2),
                                     plot.points = FALSE, 
                                     panel = function(x, y, ...) {
                                       panel.densityplot(x, ...)
                                       panel.abline(v=0.5)
                                     })
my_theme <- trellis.par.get()
my_theme$strip.background$col <- "grey80"
my_theme$plot.symbol$col <- "grey60"
my_theme$plot.polygon$col <- "grey90"

bdMHplot <- update(bdMuHist, par.settings = my_theme)
print(bdMHplot)
```

## Remarks

The results for Yule trees are equivalent for both methods. However, estimation of $\lambda$ and $\mu$ are 1) not completely equivalent as judged from the plots, and 2) more seriously, do not match the starting values in the simulations, that is, `lambda = 0.9` and `mu = 0.5`. 

This might be a limitation of the method used by `diversitree` for fitting birth-death models (said to be Nee et al., 1994), but if so we need a better way to assess with back-comparison how good are these two estimators and whether they are in fact equivalent.

Once sorted this out we can consider to use NetSim for networks in the specific case of time-trees, or more specifically, birth-death time-trees.

# Simulation spanning a fixed time interval: `sim.bdh.age`

The plot engine `NetSim::plottable.net` is said to reorient the tree for improved plotting with respect to hybrid edges. However, this also creates extinct tips for hybridization events (see the plot on the right below, sister to `t11` and compare with the same node in the plot to the left). 

This does not happen all the time, and this is the best combination of parameters that allows to reconstruct a reproducible example of such behavior. Why do this happen? Does it represent a difference in the nature of the hybridization event? Note that some other hybrid edges do not induce an "extinct" node sister to the hybrid edge.

```{r}
age_fixed <- sim.bdh.age(age = 6.7,
                         numbsim = 1,
                         lambda = 0.45,
                         mu = 0,
                         nu = 0.15,
                         hybprops = c(0.5, 0.5, 0.5),
                         hyb.inher.fxn = make.beta.draw(1,1),
                         frac = 1,
                         mrca = TRUE,
                         complete = TRUE,
                         stochsampling = FALSE,
                         hyb.rate.fxn = NULL,
                         trait.model = NULL
                         )
str(age_fixed)
par(mfrow = c(1, 2))
plot(age_fixed[[1]])
plot(plottable.net(age_fixed[[1]]))

```

\begin{small}
\begin{landscape}
\begin{longtable}[c]{p{2cm}p{1.2cm}p{1.2cm}p{1cm}p{2.6cm}p{2.6cm}p{2.6cm}p{5cm}}
\caption{Summary of arguments for simulation functions in \texttt{NetSim}. X Represents arguments present but without a default value. Filled cells otherwise represent the default values in each function, and empty cells represent arguments absent from a given function.}
\label{tab:sim_args}\\
\multicolumn{1}{c}{\textbf{argument}} & \textbf{arg.type} & \textbf{arg.str} & \textbf{arg.len} & \textbf{sim.bdh.age} & \textbf{sim.bdh.taxa.ssa} & \textbf{sim.bdh.taxa.gsa} & \multicolumn{1}{c}{\textbf{notes}} \\
\endfirsthead
%
\multicolumn{8}{c}%
{{\bfseries Table \thetable\ continued from previous page}} \\
\multicolumn{1}{c}{\textbf{argument}} & \textbf{arg.type} & \textbf{arg.str} & \textbf{arg.len} & \textbf{sim.bdh.age} & \textbf{sim.bdh.taxa.ssa} & \textbf{sim.bdh.taxa.gsa} & \multicolumn{1}{c}{\textbf{notes}} \\
\endhead
%
\hline
\texttt{age} & double & vector & 1 & X &  &  & Total simulation time length \\
\texttt{m} & integer & vector & 1 &  &  & X & Number of taxa to simulate under SSA \\
\texttt{n} & integer & vector & 1 &  & X & X & Number of final taxa \\
\texttt{numbsim} & double & vector & 1 & X & X & X & Number of networks to simulate \\
\texttt{lambda} & double & vector & 1 & X & X & X & Speciation rate \\
\texttt{mu} & double & vector & 1 & X & X & X & Extinction rate \\
\texttt{nu} & double & vector & 1 & X & X & X & Hybridization rate \\
\texttt{hybprops} & double & vector & 3 & X & X & X & Proportion of hybridizations that are lineage generative, lineage degenerative, and lineage neutral respectively \\
\texttt{hyb.inher.fxn} & closure & function & 1 & X & X & X & Function for drawing hybrid inheritance probabilities . Uses a function that draws from \texttt{rDIST}-type functions, e.g., \texttt{make.beta.draw} \\
\texttt{frac} & double & vector & 1 & \texttt{1} & \texttt{1} & \texttt{1} & Proportion of species sampled included in the phylogeny \\
\texttt{mrca} & boolean & vector & 1 & \texttt{FALSE} & \texttt{FALSE} & \texttt{FALSE} & Either origin time counts from the initial point (one lineage, \texttt{FALSE}) or from the first split (two lineages, \texttt{TRUE}) \\
\texttt{complete} & boolean & vector & 1 & \texttt{TRUE} & \texttt{TRUE} & \texttt{TRUE} & Either a tree preserving extinct lineages (if \texttt{TRUE}) or droping such extinct lineages (if \texttt{FALSE}) \\
\texttt{stochsampling} & boolean & vector & 1 & \texttt{FALSE} & \texttt{FALSE} & \texttt{FALSE} & If \texttt{TRUE}, each tip is included into the final tree with probability defined by the argument \texttt{frac}. Otherwise every lineage is guaranteed to reach the present because the phylogeny is defined as complete if \texttt{frac = 1.0} \\
\texttt{hyb.rate.fxn} & closure & function & 1 & \texttt{NULL} & \texttt{NULL} & \texttt{NULL} & Uses a function that draws from density-type functions, e.g., \texttt{make.exp.decay} \\
\texttt{trait.model} & multiple & list & 5 & \texttt{NULL} & \texttt{NULL} & \texttt{NULL} & The same structure as \texttt{make.polyploid.model}
\end{longtable}
\end{landscape}
\end{small}
